---
title: "GDP Models"
author: "Brian Howard"
date: "June 13, 2020"
output: 
  html_document:
    toc: true
    toc_depth: 4
    smart: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("getSymbols.warning4.0"=FALSE)
```

```{r libraries, echo=FALSE, message=FALSE}
library(UsingR)
library(quantmod)
library(ggplot2)
library(grid)
library(gridExtra)
library(knitr)
library(kableExtra)
library(tools)
library(zoo)
library(signal)
library(stringr)
library(corrplot)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(Quandl)
library(nnet)
library(pls)
library(readr)
#library(devtools)
#install_github("andrewuhl/RollingWindow")
library(RollingWindow)
#install.packages("EIAdata")
library(EIAdata)
library(gtable)
library(data.table)
library(readxl)
library(plotly)
```

```{r helper functions, echo=FALSE}

# Call helper functions
source("plotHelper.r")

```

```{r plot.limits, echo=FALSE}

# Define the plotting ranges
dt.recent = as.Date("2017-01-01")

# Define the S&P 500 limits
d.GSPC.max = 3500

```


```{r defsyms, echo=FALSE}

# Call code that defines the desired symbols.
source("SymbolList.r")

```



```{r load.data, echo=FALSE}

# Load data
load("C:/Users/Rainy/OneDrive//RecessionIndicator_Buffer.RData")

```


```{r symTidy, echo=FALSE}

#This snippet is needed because some of the ticker symbols include are invalid variable names so re-cast those variable names.

df.symbols$string.symbol <-
  str_replace_all(df.symbols$string.symbol, "[^[:alnum:]]", "")

```

## Feature Extraction

With the raw data downloaded, some of the interesting features can be extracted. The first step is reconcile the time intervals. Some of the data is released monthly and some daily. I chose to interpolate all data to a daily interval. The first section of code adds the daily rows to the dataframe. 

The code performs interpolation for continuous data or carries it forward for binary data like the recession indicators.

```{r aggsyms}

source("calcInterpolate.r")
df.data <- calcInterpolate(df.data, df.symbols)

```

## Create aggregate series

Some analysis requires that two or more series be combined. For example, normallizing debt by GDP to get a sense of the proportion of debt to the total economy helps understand the debt cycle.

```{r create aggregate, echo=FALSE}

source("calcAggregateSeries.r")

```

Year over year, smoothed derivative, and log trends tend to smooth out seasonal variation. It gets used so often that I do this for every series downloaded.

```{r calcsYoYSmoothLog}

source("calcFeatures.r")
lst.df <- calcFeatures(df.data, df.symbols)
df.data <- lst.df[[1]]
df.symbols <- lst.df[[2]]
```


```{r calc.features.for.aggregate, echo=FALSE}

# Calculate the features for the aggregated series
source("calcFeaturesAggregate.r")

```


```{r data.trunc, echo=FALSE}

# A very few data series do go back to 1854, but most
# don't even go past WWII so truncate the dataframe
df.data <- df.data[df.data$date > as.Date("1940-01-01"),]

```

# Recession Initiation (Switch Model)

Build the recession and recession initiation dates

```{r recframe}

source("calcRecession.r")

```

## GDP Models

### Regression

See if we can get there with a simple regression model. Begin by taking a look at how these datapoints correlate

```{r date.setup, echo=FALSE}

# Exclude the Covid19 datasets
dt.start.prediction <- as.Date(max(c(index(CLF16OV[1]), index(POPTHM[1]))))
dt.end.prediction <- as.Date("2020-01-31")

```


```{r corplot1.gdp, echo=FALSE, fig.width=12,fig.height=12}

# Correlation for the entire data set
  training.cor <- df.data[,c("GDP","INDPRO","PCE","RSALESAGG","ALTSALES")]

  rcorr.data <- rcorr(as.matrix(training.cor), type = "pearson")
  #print(rcorr.data)

  corrplot::corrplot(cor(training.cor), type="upper", order="original", 
           tl.col="black", tl.srt=45, title ="GDP Data")
```


```{r gdp.lm.setup, echo=FALSE}

df.gdp.model <-
  df.data[df.data$date >= dt.start.prediction &
            df.data$date <= dt.end.prediction,]

# I break the data into three sets: 50% for training, 25% for testing, and 25% for validation.
set.seed(123456)
in.train <- createDataPartition(y=df.gdp.model$GDP, p = 0.50, list=FALSE)
df.train <- df.gdp.model[in.train,]
df.data.rest <- df.gdp.model[-in.train,]
in.val <- createDataPartition(y = df.data.rest$GDP, p = 0.50, list = FALSE)
df.val <- df.data.rest[in.val,]
df.test <- df.data.rest[-in.val,]
rm(df.data.rest)

# Perform the linear fit
list.fit.reg <- train(
  GDP ~ INDPRO + PCE + RSALESAGG,
  data = df.train,
  method = "lm",
  preProcess = c('center', 'scale')
)


# Perform the nueral net fit
list.fit.pcr <- train(
  GDP ~ INDPRO + PCE + RSALESAGG,
  data = df.train,
  method = "pcr",
  preProcess = c('center', 'scale')
)


```

```{r gdp.lm.update, echo=FALSE}

# Add linear prediction to the dataframe
str.symbol <- "GDP.lm"
df.data[str.symbol] <-
  predict(list.fit.reg, newdata = df.data)

# Add the linear model prediction to the symbols table
df.symbols <-
  rbind(
    df.symbols,
    data.frame(
      string.symbol = str.symbol,
      string.source = "Predict",
      string.description = "Linear Model Prediction\nGDP given INDPRO,\nPCE, and RSALESAGG",
      string.label.y = getPlotYLabel(df.symbols, "POPTHM")[1],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date())
    )
  )

```

```{r gdp.pcr.update, echo=FALSE}

# Add linear prediction to the dataframe
str.symbol.pcr <- "GDP.pcr"
df.data[str.symbol.pcr] <-
  predict(list.fit.pcr, newdata = df.data)

# Add the linear model prediction to the symbols table
df.symbols <-
  rbind(
    df.symbols,
    data.frame(
      string.symbol = str.symbol.pcr,
      string.source = "Predict",
      string.description = "PCA regression\nGDP given INDPRO,\nPCE, and RSALESAGG",
      string.label.y = getPlotYLabel(df.symbols, "POPTHM")[1],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date())
    )
  )

```

```{r gdp.lm.setup.plot, echo=FALSE, fig.width = 8, fig.asp = .52}

datay <- "GDP.lm"
datay.aux <- "GDP.pcr"
datay.aux.1 <- "GDP"
ylim <- c(0, 25000)
dt.end <- Sys.Date()
myPlot <- 
  plotSingle(
    dfRecession,
    df.data,
    "date",
    datay,
    "GDP Compared to GDP predicted using linear model",
    "Date",
    getPlotYLabel(df.symbols, datay),
    c(dt.start.prediction, dt.end),
    ylim,
    b.legend = TRUE,
    b.percentile = FALSE,
    b.long.legend = TRUE
  )
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux))
  ),
  na.rm = TRUE
)
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.1))
  ),
  na.rm = TRUE
)

print(myPlot)
```

Looking at some more recent data points


```{r gdp.lm.setup.plot.recent, echo=FALSE, fig.width = 8, fig.asp = .52}

datay <- "GDP.lm"
datay.aux <- "GDP.pcr"
datay.aux.1 <- "GDP"
ylim <- c(10000, 25000)
dt.end <- Sys.Date()
dt.start <- as.Date("2018-01-01")
myPlot <-
  plotSingle(
    dfRecession,
    df.data,
    "date",
    datay,
    "GDP Compared to GDP predicted using linear model",
    "Date",
    getPlotYLabel(df.symbols, datay),
    c(dt.start, dt.end),
    ylim,
    b.legend = TRUE,
    b.percentile = FALSE,
    b.long.legend = TRUE
  )
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux))
  ),
  na.rm = TRUE
)
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.1))
  ),
  na.rm = TRUE
)

print(myPlot)
```


Look at how the GDP linear fit performed on the test data partition

```{r gdp.lm.perf, echo=FALSE}

d.test.resid = (df.test$GDP - predict(list.fit.reg, newdata = df.test))
hist(d.test.resid, main=paste("Residual For", str.symbol, sep=" "))
sd.fit.gdp.m = sd(d.test.resid)
```


```{r gdp.pcr.perf, echo=FALSE}

d.test.resid = (df.test$GDP - predict(list.fit.reg, newdata = df.test))
hist(d.test.resid, main=paste("Residual For", str.symbol.pcr, sep=" "))
sd.fit.gdp.pcr = sd(d.test.resid)

```

```{r gdp.lm.resid, echo=FALSE}

# Document the residual as well
str.symbol.res <- "GDP.lm.res"
df.data[str.symbol.res] <- df.data$GDP - df.data[str.symbol]

# Add the GDP linear model prediction residual to the symbols table
df.symbols <-
  rbind(
    df.symbols,
    data.frame(
      string.symbol = str.symbol.res,
      string.source = "Predict Resid.",
      string.description = "Residual for \nGDP given INDPRO,\nPCE, and RSALESAGG",
      string.label.y = df.symbols[df.symbols$string.symbol==str.symbol,"string.label.y"],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date())
    )
  )


```

Plot the residual  GDP to the linear fit prediction. 

The residuals decrease quite a bit during the Covid numbers, at least as of Jun 2020. Probably indicates uncertainty in the numbers.

```{r gdp.cor.resid, echo=FALSE,, fig.width = 6, fig.asp = .82}

datay <- str.symbol.res
ylim <- c(-1500, 1500)
myPlot <- plotSingle(
  dfRecession,
  df.data,
  "date",
  datay,
  getPlotTitle(df.symbols, datay),
  "Date",
  getPlotYLabel(df.symbols, datay),
  c(as.Date("1jan1967", "%d%b%Y"), Sys.Date()),
  ylim,
  b.legend = TRUE,
  b.percentile = TRUE,
  b.long.legend = FALSE
)

myPlot <- myPlot + geom_hline(yintercept=(6*sd.fit.gdp.m), color = "red", linetype="dashed")
myPlot <- myPlot + geom_hline(yintercept=(-6*sd.fit.gdp.m), color = "red", linetype="dashed")

print(myPlot)

```

