---
title: "GDP Models"
author: "Brian Howard"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    smart: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("getSymbols.warning4.0"=FALSE)
```

```{r libraries, echo=FALSE, message=FALSE}

library(tidyverse)
library(tidyquant)
library(Hmisc)
library(gridExtra)
library(signal)
library(caret)
library(data.table)

```

```{r helper functions, echo=FALSE}

# Call helper functions
source("plotHelper.r")

```


```{r load.data, echo=FALSE}

str.data.dir <- "C:/Users/Rainy/OneDrive/Documents/IssaquahDynamical/Datasets/2529RS0082_HistEconData"
out.file <- file.path(str.data.dir, "RecessionIndicator_Buffer.RData")

# Load data
load(out.file)

rm(str.data.dir)
rm(out.file)

```


```{r plot.limits, echo=FALSE}

# Define the plotting ranges
dt.recent = as.Date("2017-01-01")

# Define the start of the prediction
dt.start.prediction <- as.Date("01/02/1962", "%d/%m/%Y")

# Back test range
dt_start_back_test = as.Date('1960-01-01')

# Define the S&P 500 limits
source("SetGSPCMax.r")

```


```{r clean_data, echo=FALSE}

# Keep only the symbols needed by this code
df.symbols <- dplyr::filter(df.symbols,
                            string.symbol %in% c("USREC",
                                                 "GDP",
                                                 "GDPC1",
                                                 "GDPNOW",
                                                 "PCE",
                                                 "GDPDEF",
                                                 "PCE",
                                                 "GSG",
                                                 "XLE",
                                                 "PKG",
                                                 "IP",
                                                 "UPS",
                                                 "FDX",
                                                 "^DJI",
                                                 "INDPRO",
                                                 "RRSFS",
                                                 "RSALES",
                                                 "ALTSALES",
                                                 "MSPUS",
                                                 "HOUST"))

if( exists("b.refresh.models")){
  rm(b.refresh.models)
}

if( exists("bRefresh")){
  rm(bRefresh)
}

if( exists("d.income")){
  rm(d.income)
}

if( exists("d.Russell.max")){
  rm(d.Russell.max)
}

if( exists("dt.date")){
  rm(dt.date)
}

if( exists("dt.date.c")){
  rm(dt.date.c)
}

if( exists("i.cols")){
  rm(i.cols)
}

if( exists("idx")){
  rm(idx)
}

if( exists("list.margin.names")){
  rm(list.margin.names)
}

if( exists("list.options.names")){
  rm(list.options.names)
}

if( exists("str.name")){
  rm(str.name)
}

if( exists("str.safe.name")){
  rm(str.safe.name)
}

if( exists("str.symbol")){
  rm(str.symbol)
}

if( exists("str.symbol.raw")){
  rm(str.symbol.raw)
}

if( exists("ylimBackTest")){
  rm(ylimBackTest)
}

```

## Feature Extraction

With the raw data downloaded, some of the interesting features can be extracted.
The first step is reconcile the time intervals. Some of the data is released
monthly and some daily. I chose to interpolate all data to a daily interval. The
first section of code adds the daily rows to the dataframe.

The code performs interpolation for continuous data or carries it forward for
binary data like the recession indicators.

```{r aggsyms}

source("utilSyms.r")
source("calcInterpolate.r")
df.data <- calcInterpolate(df.symbols)

# Truncate data
df.data <- df.data[as.Date(rownames(df.data)) > as.Date("1940-01-01"),]

```

## Create aggregate series

Some analysis requires that two or more series be combined. For example, normallizing debt by GDP to get a sense of the proportion of debt to the total economy helps understand the debt cycle.

```{r create aggregate, echo=FALSE}

source("calcAggregateSeries.r")

```

Year over year, smoothed derivative, and log trends tend to smooth out seasonal variation. It gets used so often that I do this for every series downloaded.

```{r calcsYoYSmoothLog}

source("calcFeatures.r")
lst.df <- calcFeatures(df.data, df.symbols)
df.data <- lst.df[[1]]
df.symbols <- lst.df[[2]]
```


```{r calc.features.for.aggregate, echo=FALSE}

# Calculate the features for the aggregated series
source("calcFeaturesAggregate.r")

```

# Recession Initiation (Switch Model)

Build the recession and recession initiation dates

```{r recframe}

source("calcRecession.r")

```

## GDP Models

### Validate GDP Series

I was confused about what was in the GDPNow series. I created this series of plots to be sure nominal and real GDP were correct.

```{r validate_gdp_series, echo=FALSE, fig.width = 10, fig.asp = .62}

datay <- "GDP.Value__YoY"
ylim <- c(-10, 20)
dtStart = as.Date('2000-01-01')
p1 <-
  plotSingleQuick(dfRecession, df.data, datay, ylim, dtStart, b.percentile = FALSE)


datay.1 <- "GDPC1.Value__YoY"
datay.aux.1 <- "GDPNOW.Value"
ylim.1 <- c(-10, 20)
p2 <- plotSingleQuick(dfRecession, df.data, datay.1, ylim.1, dtStart, b.percentile = FALSE, b.legend = TRUE)
p2 <- p2 + geom_line(
    data = df.data,
    aes_string(x = "date",
      y = datay.aux.1,
      colour = shQuote(datay.aux.1)
    ),
    na.rm = TRUE
  )

grid.arrange(p1,
             p2,
             ncol = 1,
             top = "Nominal and real GDP")

```


```{r validate_gdp_series_yoy, echo=FALSE, fig.width = 10, fig.asp = .62}

datay <- "GDPC1.Value__YoY"
datay.aux.1 <- "GDP__by__GDPDEF__YoY"
ylim <- c(-10, 20)
dtStart = as.Date('2000-01-01')
p1 <-
  plotSingleQuick(dfRecession, df.data, datay, ylim, dtStart, b.legend <- TRUE, b.percentile = FALSE)
p1 + geom_line(
    data = df.data,
    aes_string(x = "date",
      y = datay.aux.1,
      colour = shQuote(datay.aux.1)
    ),
    na.rm = TRUE
  )

```

```{r define_ranges, echo=FALSE, fig.width = 10, fig.asp = .62}

# Define training ranges and parameters

# Exclude the Covid19 data sets
dt.start.prediction <- as.Date(max(c(index(CLF16OV[1]), index(POPTHM[1]))))
dt.end.prediction <- as.Date("2020-01-31")

```

# Supporting Series Models

The underlying data series for GDP only get updated once a month and they are
often delayed by weeks. I would like to synthesize a few series that could be
used to get more up to date estimates. I am going to start with the PCE series
since that is the largest contributor to the GDP numbers.

## PCE

### Data exploration

Truncate the data. Yahoo rarely goes past 2000

```{r pce_data_exp, echo=FALSE, fig.width = 10, fig.asp = .62}

dt.start.prediction.pce <- as.Date("2000-01-01")
df.data.pce <- df.data[as.Date(rownames(df.data)) > dt.start.prediction.pce,]

```

Starting with some correlation see if we can find market data series (updated
daily) that could be used to estimate PCE.

```{r pce_data_exp_corr, echo=FALSE, fig.width = 10, fig.asp = .62}

training.cor.pce <-
  df.data.pce[, c("PCE.Value", "GSG.Close", "XLE.Close", "PKG.Close", "IP.Close", "UPS.Close", "FDX.Close", "X_DJI.DJI.Open")]

  rcorr.data <- rcorr(as.matrix(training.cor.pce), type = "pearson")
  #print(rcorr.data)

  corrplot::corrplot(cor(training.cor.pce), type="upper", order="original", 
           tl.col="black", tl.srt=45, title ="GDP Data")

```

### Begin training up some models for PCE

```{r pce_model_dates, echo=FALSE, fig.width = 10, fig.asp = .62}

df.pce.model <-
  df.data[df.data$date >= dt.start.prediction.pce &
            df.data$date <= dt.end.prediction,]

```

#### I break the data into three sets: 50% for training, 25% for testing, and 25% for validation.

```{r pce_model_partition, echo=FALSE, fig.width = 10, fig.asp = .62}

set.seed(123456)
in.train <- createDataPartition(y = df.pce.model$PCE.Value, p = 0.50, list = FALSE)
df.train <- df.pce.model[in.train,]
df.data.rest <- df.pce.model[-in.train,]
in.val <- createDataPartition(y = df.data.rest$PCE.Value, p = 0.50, list = FALSE)
df.val <- df.data.rest[in.val,]
df.test <- df.data.rest[-in.val,]
rm(df.data.rest)

```

#### Perform the linear fit

```{r pce_model_fit_linear, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.reg.pce <- train(
  PCE.Value ~ GSG.Close + PKG.Close + UPS.Close + FDX.Close,
  data = df.train,
  method = "lm",
  preProcess = c('center', 'scale')
)

```

#### Perform the pca fit

```{r pce_model_fit_pca, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.pcr.pce <- train(
  PCE.Value ~ GSG.Close + PKG.Close + UPS.Close + FDX.Close + X_DJI.DJI.Open,
  data = df.train,
  method = "pcr",
  preProcess = c('center', 'scale')
)

```

#### Perform the knn fit

```{r pce_model_fit_knn, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.knn.pce <- train(
  PCE.Value ~ GSG.Close + PKG.Close + UPS.Close + FDX.Close + X_DJI.DJI.Open,
  data = df.train,
  method = "knn",
  preProcess = c('center', 'scale')
)

```

### Adds the linear prediction to the data frame

#### Make the prediction

```{r pce_model_sym_def, echo=FALSE, fig.width = 10, fig.asp = .62}

# Add linear prediction to the data frame
str.symbol <- "PCE.lm"
df.data[str.symbol] <-
  predict(list.fit.reg.pce, newdata = df.data)

```


```{r pce_model_sym_add, echo=FALSE, fig.width = 10, fig.asp = .62}

# Define the new description
str.description <- "Linear Model Prediction\nPCE given GSG,\nPKG, UPS, and FDX"

# Update the symbols table    
df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbol,
      string.source = "Predict",
      string.description = str.description,
      string.label.y = getPlotYLabel(df.symbols, "PCE")[1],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbol),
      string.object_name = safe_symbol_name(str.symbol)
    )
)

```

#### Create the year-over-year values

Add the linear prediction year-over-year growth

```{r pce_model_sym_yoy, echo=FALSE, fig.width = 10, fig.asp = .62}

# The year-over-year calculation for the linear model
str.symbolYoY <- paste(str.symbol, "__YoY", sep = "")
df.data[str.symbolYoY] <- CalcYoY(df.data, str.symbol, 365)

# Should be done with this
if ( exists("str.symbol")){
  rm(str.symbol)
}

# Update the symbols table    
df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbolYoY,
      string.source = "Calc",
      string.description = paste(str.description, "\nYear over Year", sep =
                                   ""),
      string.label.y = "Percent",
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbolYoY),
      string.object_name = safe_symbol_name(str.symbolYoY)
  )
)

```


### Add linear prediction by PCA to the dataframe

```{r pce_model_sym_def_pca, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol.pcr <- "PCE.pcr"
df.data[str.symbol.pcr] <-
  predict(list.fit.pcr.pce, newdata = df.data, ncomp = 1)

```

Describe the model

```{r pce_model_sym_def_pca_desc, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.pcr.pce


```

```{r pce_model_sym_add_pca, echo=FALSE, fig.width = 10, fig.asp = .62}

# Add the linear model prediction to the symbols table
str.description.pcr <- "PCE Regression Prediction\nPCE given GSG,\nPKG, UPS, and FDX"
df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbol.pcr,
      string.source = "Predict",
      string.description = str.description.pcr,
      string.label.y = getPlotYLabel(df.symbols, "PCE")[1],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbol.pcr),
      string.object_name = safe_symbol_name(str.symbol.pcr)
  )
)

```


### Add KNN prediction to the dataframe

```{r pce_model_sym_def_knn, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol.pcr <- "PCE.knn"
df.data[str.symbol.pcr] <-
  predict(list.fit.pcr.pce, newdata = df.data, ncomp = 1)

```

```{r pce_model_sym_def_knn_desc, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.knn.pce

```
Add the linear model prediction to the symbols table

```{r pce_model_sym_add_knn, echo=FALSE, fig.width = 10, fig.asp = .62}

str.description.pcr <- "KNN Prediction\nPCE given GSG,\nPKG, UPS, and FDX"
df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbol.pcr,
      string.source = "Predict",
      string.description = str.description.pcr,
      string.label.y = getPlotYLabel(df.symbols, "PCE")[1],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbol.pcr),
      string.object_name = safe_symbol_name(str.symbol.pcr)
  )
)

```

#### Add the PCA prediction year-over-year growth


```{r pce_model_sym_knn_yoy, echo=FALSE, fig.width = 10, fig.asp = .62}

# The YoY calculation
str.symbolYoY <- paste(str.symbol.pcr, "_YoY", sep = "")
df.data[str.symbolYoY] <- CalcYoY(df.data, str.symbol.pcr, 365)

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbolYoY,
      string.source = "Calc",
      string.description = paste(str.description.pcr, "\nYear over Year", sep =
                                   ""),
      string.label.y = "Percent",
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbolYoY),
      string.object_name = safe_symbol_name(str.symbolYoY)
  )
)

```


#### create PCE plot

```{r pce_model_plot, echo=FALSE, fig.width = 10, fig.asp = .62}

datay <- "PCE.Value"
datay.aux <- "PCE.lm"
datay.aux.1 <- "PCE.pcr"
datay.aux.2 <- "PCE.knn"
ylim <- c(1000, 20000)
dt.end <- Sys.Date()
myPlot <- 
  plotSingle(
    dfRecession,
    df.data,
    "date",
    datay,
    "PCE Compared to PCE predicted using linear model",
    "Date",
    getPlotYLabel(df.symbols, datay),
    c(dt.start.prediction.pce, dt.end),
    ylim,
    b.legend = TRUE,
    b.percentile = FALSE,
    b.long.legend = TRUE
  )
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux)[[1]])
  ),
  na.rm = TRUE
)
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.1))
  ),
  na.rm = TRUE
)
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.2,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.2))
  ),
  na.rm = TRUE
)


print(myPlot)

```


## GDP Models

### Regression, Retail sales

See if we can get there with a simple regression model. Begin by taking a look
at how these data points correlate

```{r gdp_corr_plot, echo=FALSE, fig.width = 10, fig.asp = .62}

training.cor <-
  df.data[, c("GDP.Value", "INDPRO.Value", "PCE.Value", "RSALESAGG", "ALTSALES.Value", "MSPUS__times__HOUST")]

  rcorr.data <- rcorr(as.matrix(training.cor), type = "pearson")
  #print(rcorr.data)

  corrplot::corrplot(cor(training.cor), type="upper", order="original", 
           tl.col="black", tl.srt=45, title ="GDP Data")

```


### Begin training up some models for GDP


```{r gdp_define_dates, echo=FALSE, fig.width = 10, fig.asp = .62}

df.gdp.model <-
  df.data[df.data$date >= dt.start.prediction &
            df.data$date <= dt.end.prediction,]

```

#### I break the data into three sets: 50% for training, 25% for testing, and 25% for validation.

```{r gdp_partition_data, echo=FALSE, fig.width = 10, fig.asp = .62}

set.seed(123456)
in.train <- createDataPartition(y=df.gdp.model$GDP.Value, p = 0.50, list=FALSE)
df.train <- df.gdp.model[in.train,]
df.data.rest <- df.gdp.model[-in.train,]
in.val <- createDataPartition(y = df.data.rest$GDP.Value, p = 0.50, list = FALSE)
df.val <- df.data.rest[in.val,]
df.test <- df.data.rest[-in.val,]
rm(df.data.rest)

```

#### Perform the linear fit

```{r gdp_fit_lm, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.reg <- train(
  GDP.Value ~ INDPRO.Value + PCE.Value + RSALESAGG + ALTSALES.Value + MSPUS__times__HOUST,
  data = df.train,
  method = "lm",
  preProcess = c('center', 'scale')
)

```

#### Perform the pca fit

```{r gdp_fit_pca, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.pcr <- train(
  GDP.Value ~ INDPRO.Value + PCE.Value + RSALESAGG + ALTSALES.Value + MSPUS__times__HOUST,
  data = df.train,
  method = "pcr",
  preProcess = c('center', 'scale')
)

```

### Adds the linear prediction to the data frame

Add the nominal linear prediction year-over-year growth

#### Add linear prediction to the dataframe

```{r gdp_fit_ln_add, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol <- "GDP.lm"
df.data[str.symbol] <-
  predict(list.fit.reg, newdata = df.data)

# Add the linear model prediction to the symbols table
str.description <- "Linear Model Prediction\nGDP given INDPRO,\nPCE, RSALESAGG, ALTSALES,\nand housing"

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbol,
      string.source = "Predict",
      string.description = str.description,
      string.label.y = getPlotYLabel(df.symbols, "POPTHM")[1],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbol),
      string.object_name = safe_symbol_name(str.symbol)
  )
)

```

#### Make the year over year calculation

```{r gdp_fit_lm_norm_yoy_calc, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbolYoY <- paste(str.symbol, "_YoY", sep = "")
df.data[str.symbolYoY] <- CalcYoY(df.data, str.symbol, 365)

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbolYoY,
      string.source = "Calc",
      string.description = paste(str.description, "\nYear over Year", sep =
                                   ""),
      string.label.y = "Percent",
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbolYoY),
      string.object_name = safe_symbol_name(str.symbolYoY)
  )
)

# Clean and tidy memory  
if( exists("str.symbolYoY")){
  rm(str.symbolYoY)
}
 
if( exists("str.symbol")){
  rm(str.symbol)
}

```

#### Normalize linear model GDP commodities by GDP deflator

```{r gdp_fit_ln_norm_def, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol <- "GDP.lm.by.GDPDEF" 
df.data[str.symbol] <-
  (df.data$GDP.lm / df.data$GDPDEF.Value)

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = "GDP.lm.by.GDPDEF",
      string.source = "Calc",
      string.description = "Linear Model Est. GDP \nNormalized by GDP def",
      string.label.y = "(-)",
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start =  as.Date(max(c(
        index(GDP[1]), index(GDPDEF[1])
      ))) ,
      date.series.end = as.Date(min(c(
        index(tail(GDP, 1)), index(tail(GDPDEF, 1))
      ))),
      string.symbol_safe = safe_symbol_name(str.symbol),
      string.object_name = safe_symbol_name(str.symbol)
  )
)

```


Add the real linear prediction (nominal/deflator) year-over-year growth

```{r gdp_fit_ln_norm_lm_def, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol <- "GDP.lm.by.GDPDEF"
str.symbolYoY <- paste(str.symbol, "_YoY", sep = "")
df.data[str.symbolYoY] <- CalcYoY(df.data, str.symbol, 365)

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbolYoY,
      string.source = "Calc",
      string.description = paste(str.description, "\nYear over Year", sep =
                                   ""),
      string.label.y = "Percent",
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbolYoY),
      string.object_name = safe_symbol_name(str.symbolYoY)
  )
)

```

### Add linear prediction by PCA to the dataframe

```{r gdp_pred_pcr_calc, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol.pcr <- "GDP.pcr"
df.data[str.symbol.pcr] <-
  predict(list.fit.pcr, newdata = df.data, ncomp = 1)


```


```{r gdp_pred_pcr_describe, echo=FALSE, fig.width = 10, fig.asp = .62}


list.fit.pcr  

```

#### Add the linear model prediction to the symbols table

```{r gdp_pred_add_to_syms, echo=FALSE, fig.width = 10, fig.asp = .62}

str.description.pcr <- "PCA regression\nGDP given INDPRO,\nPCE, RSALESAGG, ALTSALES,\nand housing"  

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbol.pcr,
      string.source = "Predict",
      string.description = str.description.pcr,
      string.label.y = getPlotYLabel(df.symbols, "PCE")[1],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date())
  )
)

```

Add the PCA prediction year-over-year growth

```{r gdp_pred_pcr_yoy, echo=FALSE, fig.width = 10, fig.asp = .62}

# The YoY calculation for PCA
str.symbolYoY <- paste(str.symbol.pcr, "_YoY", sep = "")
df.data[str.symbolYoY] <- CalcYoY(df.data, str.symbol.pcr, 365)

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbolYoY,
      string.source = "Calc",
      string.description = paste(str.description.pcr, "\nYear over Year", sep =
                                   ""),
      string.label.y = "Percent",
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date())
  )
)

# Clean and tidy memory  
if( exists("str.description.pcr")){
  rm(str.description.pcr)
}
```

Plot the predicted and actual values

```{r gdp_pred_plot_long, echo=FALSE, fig.width = 10, fig.asp = .62}

datay <- "GDP.lm"
datay.aux <- "GDP.pcr"
datay.aux.1 <- "GDP.Value"
ylim <- c(0, 35000)
dt.end <- Sys.Date()
myPlot <- 
  plotSingle(
    dfRecession,
    df.data,
    "date",
    datay,
    "GDP Compared to GDP predicted using linear model",
    "Date",
    getPlotYLabel(df.symbols, datay),
    c(dt.start.prediction, dt.end),
    ylim,
    b.legend = TRUE,
    b.percentile = FALSE,
    b.long.legend = TRUE
  )
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux))
  ),
  na.rm = TRUE
)
myPlot <- myPlot + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.1))
  ),
  na.rm = TRUE
)

print(myPlot)

```
  
  
Looking at some more recent data points

```{r gdp_pred_plot_near, echo=FALSE, fig.width = 10, fig.asp = .62}

lst_syms <- c("GDP.Value",
              "GDP.lm",
              "GDP.pcr")
if ( require_columns(df.data, lst_syms ) ){
    
  dt.end <- Sys.Date()
  dt.start <- as.Date("2015-01-01")
  myPlot <-
    plotSingle(
      dfRecession,
      df.data,
      "date",
      datay = lst_syms[[1]],
      "GDP Compared to GDP predicted values",
      "Date",
      getPlotYLabel(df.symbols, lst_syms[[1]]),
      c(dt.start, dt.end),
      ylim =  c(10000, round_up_to(max(df.data[[lst_syms[[1]]]]),1)),
      b.legend = TRUE,
      b.percentile = FALSE,
      b.long.legend = TRUE
    )
  myPlot <- myPlot + geom_line(
    data = df.data,
    aes(
      x = .data[["date"]],
      y = .data[[lst_syms[[2]]]],
      colour = getPlotTitle(df.symbols, lst_syms[[2]])
    ),
    na.rm = TRUE
  )
  myPlot <- myPlot + geom_line(
    data = df.data,
    aes(
      x = .data[["date"]],
      y = .data[[lst_syms[[3]]]],
      colour = getPlotTitle(df.symbols, lst_syms[[3]])
    ),
    na.rm = TRUE
  )
  
  print(myPlot)

  # Clean and tidy memory
  if( exists("my_plot")){
    rm(my_plot)  
  }
  
}

# Clean and tidy memory
if( exists("lst_syms")){
  rm(lst_syms)
}

```  

```{r gdp_pred_plot_long_yoy, echo=FALSE, fig.width = 10, fig.asp = .62}

datay <- "GDP.Value__YoY"
datay.aux <- "GDP.lm_YoY"
datay.aux.1 <- "GDP.pcr_YoY"
ylim <- c(-50, 50)
dt.end <- Sys.Date()
dt.start <- as.Date("2015-01-01")
p1 <-
  plotSingle(
    dfRecession,
    df.data,
    "date",
    datay,
    "GDP Compared to GDP predicted using linear model",
    "Date",
    getPlotYLabel(df.symbols, datay),
    c(dt.start, dt.end),
    ylim,
    b.legend = TRUE,
    b.percentile = FALSE,
    b.long.legend = TRUE
  )
p1 <- p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux))
  ),
  na.rm = TRUE
)
p1 <- p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.1))
  ),
  na.rm = TRUE
)

datay.1 <- "GDP.lm.by.GDPDEF_YoY"
datay.aux.1 <- "GDPNOW.Value"
ylim.1 <- c(-10, 20)
p2 <- plotSingleQuick(dfRecession, 
                      df.data, datay.1, 
                      ylim.1, dt.start = c(dt.start, dt.end), 
                      b.percentile = FALSE, 
                      b.legend = TRUE,
                      b.long.legend = TRUE)
p2 <- p2 + geom_line(
    data = df.data,
    aes_string(x = "date",
      y = datay.aux.1,
      colour = shQuote(datay.aux.1)
    ),
    na.rm = TRUE
  )

grid.arrange(p1,
             p2,
             ncol = 1,
             top = "Nominal and real GDP")  

```

Look at how the GDP linear fit performed on the test data partition

```{r gdp_lm_resid_calc, echo=FALSE, fig.width = 10, fig.asp = .62}


d.test.resid = (df.test$GDP.Value - predict(list.fit.reg, newdata = df.test))
hist(d.test.resid, main=paste("Residual For", str.symbol, sep=" "))
sd.fit.gdp.m = sd(d.test.resid)

```

Print out the coefficients to see how the linear weights the input values

```{r gdp_lm_resid_calc_desc, echo=FALSE, fig.width = 10, fig.asp = .62}

list.fit.reg$finalModel

```


```{r gdp_pcr_resid_calc, echo=FALSE, fig.width = 10, fig.asp = .62}

d.test.resid = (df.test$GDP.Value - predict(list.fit.pcr, newdata = df.test))
hist(d.test.resid, main=paste("Residual For", str.symbol.pcr, sep=" "))
sd.fit.gdp.pcr = sd(d.test.resid)

```

```{r gdp_pcr_resid_calc_desc, echo=FALSE, fig.width = 10, fig.asp = .62}

summary(list.fit.pcr)

```

Document the residual as well


```{r gdp_lm_resid_calc_syms, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol.res <- "GDP.lm.res"
df.data[str.symbol.res] <- df.data$GDP.Value - df.data$GDP.lm

# Add the GDP linear model prediction residual to the symbols table
df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbol.res,
      string.source = "Predict Resid.",
      string.description = "Residual for \nGDP given INDPRO,PCE, RSALESAGG,\nauto sales, and home sales",
      string.label.y = df.symbols[df.symbols$string.symbol==str.symbol,"string.label.y"],
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbolYoY),
      string.object_name = safe_symbol_name(str.symbolYoY)
  )
)

# Clean and tidy memory  
if( exists("str.symbol.res")){
  rm(str.symbol.res)
}

```

Document the PCR residual


```{r gdp_pcr_resid_calc_syms, echo=FALSE, fig.width = 10, fig.asp = .62}


str.symbol.pcr.res <- "GDP.pcr.res"
df.data[str.symbol.pcr.res] <- df.data$GDP.Value - df.data$GDP.pcr

# Add the GDP linear model prediction residual to the symbols table

df.symbols <- symbols_append_row(
    df.symbols,
    list(
      string.symbol = str.symbol.pcr.res,
      string.source = "Predict Resid.",
      string.description = "PCR Residual for \nGDP given INDPRO,\nPCE, and RSALESAGG",
      string.label.y = getPlotYLabel(df.symbols, "GDP"),
      float.expense.ratio = -1.00,
      Max030 = FALSE,
      Max180 = FALSE,
      date.series.start = dt.start.prediction,
      date.series.end = as.Date(Sys.Date()),
      string.symbol_safe = safe_symbol_name(str.symbolYoY),
      string.object_name = safe_symbol_name(str.symbolYoY)
  )
)

# Clean and tidy memory  
if( exists("str.symbol.pcr.res")){
  rm(str.symbol.pcr.res)
}

```


Plot the residual  GDP to the linear fit prediction. 

The residuals decrease quite a bit during the Covid numbers, at least as of Jun 2020. Probably indicates uncertainty in the numbers.

```{r gdp_lm_resid_calc_plot, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol.res <- "GDP.lm.res"
ylim <- c(-2000, 2000)
myPlot <- plotSingle(
  dfRecession,
  df.data,
  "date",
  datay = str.symbol.res,
  getPlotTitle(df.symbols, str.symbol.res),
  "Date",
  getPlotYLabel(df.symbols, str.symbol.res),
  c(as.Date("1jan1967", "%d%b%Y"), Sys.Date()),
  ylim,
  b.legend = TRUE,
  b.percentile = TRUE,
  b.long.legend = FALSE
)

myPlot <- myPlot + geom_hline(yintercept=(6*sd.fit.gdp.m), color = "red", linetype="dashed")
myPlot <- myPlot + geom_hline(yintercept=(-6*sd.fit.gdp.m), color = "red", linetype="dashed")

print(myPlot)

# Clean and tidy memory  
if( exists("str.symbol.res")){
  rm(str.symbol.res)
}

if( exists("myPlot")){
  rm(myPlot)
}

```

Plot the residual GDP to the PCR fit prediction. 

The residuals decrease quite a bit during the Covid numbers, at least as of Jun
2020. Probably indicates uncertainty in the numbers.

```{r gdp_pcr_resid_calc_plot, echo=FALSE, fig.width = 10, fig.asp = .62}

str.symbol <- "GDP.pcr.res"
ylim <- c(-9000, 9000)
myPlot <- plotSingle(
  dfRecession,
  df.data,
  "date",
  datay = str.symbol,
  getPlotTitle(df.symbols, str.symbol),
  "Date",
  getPlotYLabel(df.symbols, str.symbol),
  c(as.Date("1jan1967", "%d%b%Y"), Sys.Date()),
  ylim,
  b.legend = TRUE,
  b.percentile = TRUE,
  b.long.legend = FALSE
)

myPlot <- myPlot + geom_hline(yintercept=(6*sd.fit.gdp.pcr), color = "red", linetype="dashed")
myPlot <- myPlot + geom_hline(yintercept=(-6*sd.fit.gdp.pcr), color = "red", linetype="dashed")

print(myPlot)

# Clean and tidy memory  
if( exists("str.symbol.res")){
  rm(str.symbol.res)
}

if( exists("myPlot")){
  rm(myPlot)
}

```




