---
title: "Unemployment Models"
author: "Brian Howard"
date: "July 2, 2020"
output: 
  html_document:
    toc: true
    toc_depth: 4
    smart: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("getSymbols.warning4.0"=FALSE)
```

```{r libraries, echo=FALSE, message=FALSE}
library(UsingR)
library(quantmod)
library(ggplot2)
library(grid)
library(gridExtra)
library(knitr)
library(kableExtra)
library(tools)
library(zoo)
library(signal)
library(stringr)
library(corrplot)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(Quandl)
library(nnet)
library(readr)
#library(devtools)
#install_github("andrewuhl/RollingWindow")
library(RollingWindow)
#install.packages("EIAdata")
library(EIAdata)
library(gtable)
library(data.table)
library(readxl)
library(plotly)
```

```{r helper functions, echo=FALSE}

# Call helper functions
source("plotHelper.r")

```

```{r plot.limits, echo=FALSE}

# Define the plotting ranges
dt.recent = as.Date("2017-01-01")

# Define the S&P 500 limits
d.GSPC.max = 3500

```


```{r defsyms, echo=FALSE}

# Call code that defines the desired symbols.
source("SymbolList.r")

```



```{r load.data, echo=FALSE}

# Load data
load("C:/Users/Rainy/OneDrive//RecessionIndicator_Buffer.RData")

```

```{r symTidy, echo=FALSE}

#This snippet is needed because some of the ticker symbols include are invalid variable names so re-cast those variable names.

df.symbols$string.symbol <-
  str_replace_all(df.symbols$string.symbol, "[^[:alnum:]]", "")

```

## Feature Extraction

With the raw data downloaded, some of the interesting features can be extracted. The first step is reconcile the time intervals. Some of the data is released monthly and some daily. I chose to interpolate all data to a daily interval. The first section of code adds the daily rows to the dataframe. 

The code performs interpolation for continuous data or carries it forward for binary data like the recession indicators.

```{r aggsyms}

source("calcInterpolate.r")
df.data <- calcInterpolate(df.data, df.symbols)

```

## Truncate data

```{r data.trunc, echo=FALSE}

# A very few data series do go back to 1854, but most
# don't even go past WWII so truncate the dataframe
df.data1 <- df.data[as.Date(rownames(df.data)) > as.Date("1940-01-01"),]

```

## Create aggregate series

Some analysis requires that two or more series be combined. For example, normallizing debt by GDP to get a sense of the proportion of debt to the total economy helps understand the debt cycle.

```{r create aggregate, echo=FALSE}

source("calcAggregateSeries.r")

```

Year over year, smoothed derivative, and log trends tend to smooth out seasonal variation. It gets used so often that I do this for every series downloaded.

```{r calcsYoYSmoothLog}

source("calcFeatures.r")
lst.df <- calcFeatures(df.data, df.symbols)
df.data <- lst.df[[1]]
df.symbols <- lst.df[[2]]
```


```{r calc.features.for.aggregate, echo=FALSE}

# Calculate the features for the aggregated series
source("calcFeaturesAggregate.r")

```


# Recession initiation and termination dates/times

Build the recession and recession initiation dates

```{r recframe}

source("calcRecession.r")

```

## Historical statistics

### Year over year

How much did the market change in a year, or 4, or 5? This data is used to evaluate the probability of a XX% decline in a certain period.

Start with year-over-year

```{r GSPC.Close_YoY1, echo=FALSE, echo=FALSE, fig.width = 8, fig.asp = 0.8  }
datay.in <- "GSPC.Close_YoY"
datay.aux.1 <- "DJI.Close_YoY"
datay.aux.2 <- "RLG.Close_YoY"
ylim.in <- c(-75, 75)
dt.start <- as.Date('1940-01-01')
p1 <- plotSingle(datadf_rec=dfRecession,
           datadf=df.data,
           datax="date",
           datay = datay.in,
           titlelabel = getPlotTitle(df.symbols, datay.in),
           xlabel = "Date",
           ylabel = getPlotYLabel(df.symbols, datay.in),
           xlim = c(dt.start, Sys.Date()),
           ylim = ylim.in,
           b.legend = TRUE,
           b.percentile = TRUE,
           b.long.legend = TRUE) 
p1 <-p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(datay.aux.1)
  ),
  na.rm = TRUE
)

p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.2,
    colour = shQuote(datay.aux.2)
  ),
  na.rm = TRUE
)

```

See how it looks over a 4-year period

```{r GSPC.Close_YoY4, echo=FALSE, fig.width = 8, fig.asp = 0.8  }
datay.in <- "GSPC.Close_YoY4"
datay.aux.1 <- "DJI.Close_YoY4"
datay.aux.2 <- "RLG.Close_YoY4"
ylim.in <- c(-100, 200)
dt.start <- as.Date('1940-01-01')
p1 <- plotSingle(datadf_rec=dfRecession,
           datadf=df.data,
           datax="date",
           datay = datay.in,
           titlelabel = "4 Year - 4 Year change, GSPC, DJI, RLS",
           xlabel = "Date",
           ylabel = getPlotYLabel(df.symbols, datay.in),
           xlim = c(dt.start, Sys.Date()),
           ylim = ylim.in,
           b.legend = TRUE,
           b.percentile = TRUE,
           b.long.legend = TRUE) 
p1 <-p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.1))
  ),
  na.rm = TRUE
)

p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.2,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.2))
  ),
  na.rm = TRUE
)
```

With a 4-year window, what are the probabilities of certain returns?

```{r GSPC.Close_YoY4.hist, echo=FALSE, fig.width = 8, fig.asp = 0.8  }

data.hist <- df.data$GSPC.Close_YoY4[df.data$date>dt.start]
my.h <- hist(data.hist,breaks=500,main="S&P 500 4 year returns")
abline(v = median(data.hist), col = "blue", lwd = 2)
my.h$counts=100*my.h$counts/sum(my.h$counts)
plot(my.h)
abline(v = median(data.hist), col = "blue", lwd = 2)

```


See how it looks over a 5-year period

```{r GSPC.Close_YoY5, echo=FALSE, fig.width = 8, fig.asp = 0.8  }
datay.in <- "GSPC.Close_YoY5"
datay.aux.1 <- "DJI.Close_YoY5"
datay.aux.2 <- "RLG.Close_YoY5"
ylim.in <- c(-100, 300)
dt.start <- as.Date('1930-01-01')
p1 <- plotSingle(datadf_rec=dfRecession,
           datadf=df.data,
           datax="date",
           datay = datay.in,
           titlelabel = "5 Year - 5 Year change, GSPC, DJI, RLS",
           xlabel = "Date",
           ylabel = getPlotYLabel(df.symbols, datay.in),
           xlim = c(dt.start, Sys.Date()),
           ylim = ylim.in,
           b.legend = TRUE,
           b.percentile = TRUE,
           b.long.legend = TRUE) 
p1 <-p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.1,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.1))
  ),
  na.rm = TRUE
)

p1 + geom_line(
  data = df.data,
  aes_string(
    x = "date",
    y = datay.aux.2,
    colour = shQuote(getPlotTitle(df.symbols, datay.aux.2))
  ),
  na.rm = TRUE
)
```

With a 5-year window, what are the probabilities of certain returns?

```{r GSPC.Close_YoY5.hist, echo=FALSE, fig.width = 8, fig.asp = 0.8  }

data.hist <- df.data$GSPC.Close_YoY5[df.data$date>dt.start]
my.h <- hist(data.hist,breaks=500,main="S&P 500 5 year returns")
abline(v = median(data.hist), col = "blue", lwd = 2)
my.h$counts=100*my.h$counts/sum(my.h$counts)
plot(my.h)
abline(v = median(data.hist), col = "blue", lwd = 2)


```


