---
title: "Portfolio Testing"
author: "Brian Howard"
date: "June 27, 2017"
output:
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
---

# Function Library

This section contains the functions used in the code.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("getSymbols.warning4.0"=FALSE)
strDateStart <- "2011-01-01"
```

```{r runopts, echo=FALSE}

# Set this to true to download new data. Otherwise it pulls it in from the disk.
bDownload = TRUE
iRetPd <- 365

```

```{r libraries, echo=FALSE, message=FALSE}
library(knitr)
library(quantmod)
library(zoo)
library(stringr)
library(ggplot2)
library(data.table)
library(corrplot)
library(Hmisc)
library(tseries)
```


```{r porthelp, echo=FALSE}

pfUpdateReturn <- function(strPortfolio, dfData, dfSyms){

  dfData[,strPortfolio] <- 0
  for (col_name in names(dfData))
  {
      # Act only the numeric data
      if (is.numeric(dfData[,col_name]))
      {
        
        # Split the name ("USGFG.Close"" is "USGFG"" and "Close"")
        lstSyms <- lstSymSplit(col_name)
        
        # Only if there is two terms
        if( length(lstSyms) > 1){
  
          if( lstSyms[2] == 'Close_Norm' ){
  
            dPercent <- dfSyms[dfSyms$Symbol==lstSyms[1],strPortfolio]
            if(length(dPercent) > 0 ){
              dfData[,strPortfolio] <- dfData[,strPortfolio] + (dfData[,col_name] * dPercent)
            }
          }
                  
        }
        
      }
  }
  
  return(dfData)
  
}

```

```{r funchelp, echo=FALSE}

strSymOnly <- function(datay){
  
  return(lstSymSplit(datay)[1])
}

lstSymSplit <- function(datay){
  
  lstSyms <- unlist(strsplit(datay, "\\."))
  return(lstSyms)
}


```

```{r plothelp }
# ------------------------------------------------------------------------------
# Define the generic single plane plot function
plotSingle <- function(datadf, datax, datay, titlelabel, xlabel, ylabel, xlim, ylim, bLegend){
  
  myplot <- ggplot()+
    theme(plot.background = element_rect(fill = 'white', colour = 'white')) +
    theme(panel.background = element_rect(fill = 'white', colour = 'grey')) +
    theme(panel.grid.major.x = element_blank()) +
    theme(panel.grid.major.y = element_line(colour="grey", size=0.5)) +
    theme(panel.grid.minor.y = element_line(colour="grey", size=0.25, linetype = "dotted")) +
    geom_line(data=datadf, aes_string(x=datax, y=datay, colour = factor(datay)), 
              na.rm = TRUE, size = 1.0) +
    scale_colour_manual(values = c("black","blue", "green")) +
    guides(colour = guide_legend("Series"), size = guide_legend("Series"), shape = guide_legend("Series")) +
    scale_fill_continuous(name = "V") +
    #geom_smooth(method = "lm") + 
    ggtitle(titlelabel) +
    labs(x=xlabel, y = ylabel) +
    scale_x_date(limits = xlim ) + 
    scale_y_continuous(limits = ylim) +
    if( bLegend){
      theme(legend.position = "top")
    }else{
      theme(legend.position = "none")
    }
  
  

  return(myplot)
}

# ------------------------------------------------------------------------------
# Define the function for calculating year over year growth. 
CalcYoY <- function (datadf, strCol, iPeriods){
  Nrow <- nrow(datadf)
  GrowthRateYoY <- rep(0,Nrow)
  GrowthRateYoY[(iPeriods+1):Nrow] <- diff(as.matrix(datadf[[strCol]]), lag = iPeriods)
  GrowthRateYoY <- (GrowthRateYoY / datadf[[strCol]])
  return(GrowthRateYoY)
}

# Define return using log
CalcLogRet <- function(datadf, strCol, iPeriods){
  Nrow <- nrow(datadf)
  GrowthRate <- rep(0,Nrow)
  GrowthRate[(iPeriods+1):Nrow] <- diff(log(datadf[[strCol]]), lag = iPeriods)
  return(GrowthRate)
}


# Small helper function to get the symbol description
getPlotTitle <- function(datay){
  strTitle <-  paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$Desc)
  return(strTitle)
}

getPlotYLabel <- function(datay){
  strY <- dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$yLabel
  return(strY)
}

plotSingleBench <- function(datay, ylim){
  
  if( grep('.',datay) > 0 ){
    strSym <- strSymOnly(datay)
    strTitle <- paste(datay, " | ", dfSyms[grep(strSym, dfSyms$Symbol),]$Desc)
    strYLabel <- dfSyms[grep(strSym, dfSyms$Symbol),]$yLabel
  }else{
    strTitle <- paste(datay, " | ", dfSyms[grep(datay, dfSyms$Symbol),]$Desc)
    strYLabel <- dfSyms[grep(datay, dfSyms$Symbol),]$yLabel
  }
  dataBench <- "GSPC.Close_Norm"
  datax <- "date"
  myPlot <- plotSingle(dfData, datax, datay, 
            strTitle, "Date", 
            strYLabel, c(as.Date(strDateStart), Sys.Date()), ylim, TRUE)
  myPlot <- myPlot + geom_line(data=dfData, aes_string(x=datax, y=dataBench, colour = factor(dataBench)), 
          na.rm = TRUE, size = 0.7)
  
  
  return(myPlot)
 
  
}

plotSingleQuick <- function(datay, ylim){
  
  if( grep('.',datay) > 0 ){
    strSym <- strSymOnly(datay)
    strTitle <- paste(datay, " | ", dfSyms[grep(strSym, dfSyms$Symbol),]$Desc)
    strYLabel <- dfSyms[grep(strSym, dfSyms$Symbol),]$yLabel
  }else{
    strTitle <- paste(datay, " | ", dfSyms[grep(datay, dfSyms$Symbol),]$Desc)
    strYLabel <- dfSyms[grep(datay, dfSyms$Symbol),]$yLabel
  }
  datax <- "date"
  myPlot <- plotSingle(dfData, datax, datay, 
            strTitle, "Date", 
            strYLabel, c(as.Date(strDateStart ), Sys.Date()), ylim, TRUE)
  
  return(myPlot)
 
  
}

plotRR <- function(datay, bIgnorePort){
  
  # Handle optional argurements
  if( missing(bIgnorePort)){
    bIgnorePort = FALSE;
  }

  # Create a new data frame for just this portfolio
  dfRR <- dfSyms[dfSyms[datay]>0,]
  
  # Need a data table with just the ticker and data
  strCols <- paste(dfRR$Symbol, ".Close_Norm_YoY", sep="")
  dfPort <- data.table(dfData[,strCols])

  # Range of expected returns from the porfolio
  er_vals <- seq(from = min(dfRR$ExpReturn), to = max(dfRR$ExpReturn), length.out = 1000)
  
  # find an optimal portfolio for each possible possible expected return 
  # (note that the values are explicitly set between the minimum and maximum of the expected returns per asset)
  sd_vals <- rep(0,length(er_vals))
  tryCatch({
    sd_vals <- sapply(er_vals, function(er) {
      op <- portfolio.optim(as.matrix(dfPort), er)
      return(op$ps)
    })
  }, error = function(e) {
    print("Failed to find efficiency boundary")
  })
  
  # Collect in a table
  plot_dt <- data.table(sd = sd_vals, er = er_vals)
  
  # find the lower and the upper frontier
  minsd <- min(plot_dt$sd)
  minsd_er <- plot_dt[sd == minsd, er]
  minsd_er <- minsd_er[1]
  plot_dt[, efficient := er >= minsd_er]
  
  # Data for the actual portfolio mix
  dfPortRet <- dfData[datay]
  strPfYoY <- paste(datay,"_YoY",sep="") 
  dfPortRet[strPfYoY] <- CalcYoY(dfPortRet, datay, iRetPd)
  Vol_pf <- sd(dfPortRet[,strPfYoY])
  ExpRet_pf <- mean(dfPortRet[,strPfYoY])
  #print(Vol_pf)
  #print(ExpRet_pf)
  
  # Data for the portfolio mix with same returns, but lower volatility
  dfRR[,paste(datay,"_Opt",sep="")] <- 0
  op_pf <- data.frame(0)
  op_pf$ps <- 0
  tryCatch({
      op_pf <- portfolio.optim(as.matrix(dfPort), ExpRet_pf)
      dfRR[paste(datay,"_Opt",sep="")] <- op_pf$pw

    }, error=function(e) {
      print("Failed to find opt portfolio")
    })

# Plot the data
  myPlot <- ggplot() + 
    geom_point(data=dfRR, aes(x = Volatility, y = ExpReturn, size=5, color = Symbol)) +
    geom_point(data = plot_dt[efficient == F], aes(x = sd, y = er), size = 0.5, color = "blue") +
    geom_point(data = plot_dt[efficient == T], aes(x = sd, y = er), size = 0.5, color = "red") +
    theme_bw() + ggtitle("Risk-Return Tradeoff (Red=Efficient), Annual Returns") + 
    xlab("Volatility") + ylab("Expected Returns") +
    scale_y_continuous(label = scales::percent, limits = c(-0.05, 0.15)) +
    scale_x_continuous(label = scales::percent, limits = c(0.00, 0.30))

    if( bIgnorePort==FALSE ){
      myPlot <- myPlot + geom_point(aes(x = Vol_pf, y = ExpRet_pf, size=5, color = "Actual Portfolio")) +
      geom_point(aes(x = op_pf$ps, y = ExpRet_pf, size=5, color = "Reduced Risk Portfolio"))
    }
  
  return(list(dfRR, myPlot, plot_dt))

}

plotCorr <- function(strPort){
  
    lstActiveSyms <- paste(dfSyms[dfSyms[,strPort]>0,]$Symbol,'.Close_Norm',sep="")


  # Correlation for the entire data set
  training.cor <- dfData[,lstActiveSyms]
  rcorr.data <- rcorr(as.matrix(training.cor), type = "pearson")
  #print(rcorr.data)

  corrplot(cor(training.cor), type="upper", order="original", 
           tl.col="black", tl.srt=45, title ="All data")
}


```

These functions help organize tables in the document

```{r tabhelp}

tblPort <- function(rrData, datay){
  
  kable(rrData[[1]][c("Symbol", "Desc", datay, paste(datay, "_Opt", sep = ""))])
  
}

```

# Introduction

This is a portfolio analysis, data pulled from yahoo. I'm trying to decide what a better mix would be for this old IRA account. These next few segments load the data into the R program.

## Define the symbols to be used in the analysis.

This code pulls in the library stock ticker symbols that will be available for analysis.

```{r defsyms, echo=FALSE}
Symbol = c("USPRX", "USSPX")
Source = c("yahoo", "yahoo")
Desc = c("USAA S&P 500 Index Reward",
         "USAA S&P 500 Index Member")
yLabel = c("Dollars","Dollars")
dfSyms = data.frame(Symbol, Source, Desc, yLabel, stringsAsFactors = FALSE)

dfSyms <- rbind(dfSyms, data.frame(Symbol="USAWX", Source="yahoo", 
              Desc="USAA World Growth", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="USNQX", Source="yahoo", 
              Desc="USAA NASDAQ-100 Index", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="USISX", Source="yahoo", 
              Desc="USAA Income Stock", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="UFSGX", Source="yahoo", 
              Desc="USAA First Start Growth", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="USCGX", Source="yahoo", 
              Desc="USAA Capital Growth Fund", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="USCCX", Source="yahoo", 
              Desc="Cornerstone Conservative Fund", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="USAAX", Source="yahoo", 
              Desc="Growth Fund", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="VTWO", Source="yahoo", 
              Desc="Vanguard Russell 2000 ETF", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="VFINX", Source="yahoo", 
              Desc="Vanguard 500 Index Investor", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="VGSTX", Source="yahoo", 
              Desc="Vanguard STAR Inv", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="TMFGX", Source="yahoo", 
              Desc="Motley Fool Great America Investor", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="IWM", Source="yahoo", 
              Desc="iShares Russell 2000", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="QQQ", Source="yahoo", 
              Desc="PowerShares QQQ ETF", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="HAINX", Source="yahoo", 
              Desc="Harbor International Institutional", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="VEU", Source="yahoo", 
              Desc="Vanguard FTSE All-Wld ex-US ETF", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="IVOO", Source="yahoo", 
              Desc="Vanguard S&P Mid-Cap 400 ETF", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="SPY", Source="yahoo", 
              Desc="SPDR S&P 500 ETF", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="VIGI", Source="yahoo", 
              Desc="Vanguard Whitehall Funds - Vanguard Int. Div.", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="CAMIX", Source="yahoo", 
              Desc="Cambiar International Equity Inv", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="SCHD", Source="yahoo", 
              Desc="Schwab U.S. Dividend Equity ETF", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="VYM", Source="yahoo", 
              Desc="Vanguard High Dividend Yield ETF", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="NOBL", Source="yahoo", 
              Desc="ProShares S&P 500 Dividend Aristocrats", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="XOM", Source="yahoo", 
              Desc="Exxon Mobil Corporation", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="MSFT", Source="yahoo", 
              Desc="Microsoft Corporation", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="JNJ", Source="yahoo", 
              Desc="Johnson & Johnson", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="INTC", Source="yahoo", 
              Desc="Intel Corporation", 
              yLabel="Dollars" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="^GSPC", Source="yahoo", 
              Desc="S&P 500",
              yLabel = "Dollars" ) )

```

## Build the data frame

This code loops through the dfSyms dataframe and pulls down the symbols. The getSymbols function returns zoo objects, this snippet converts it into a dataframe, which simplifies operations in R.

```{r getsyms, echo=FALSE}

if( bDownload ){

  # Get the data
  for (idx in 1:nrow(dfSyms)){
    getSymbols(as.character(dfSyms[idx,"Symbol"]),
               src=as.character(dfSyms[idx,"Source"]), 
               auto.assign=TRUE, 
               from = as.Date("1900-01-01"), to = Sys.Date())
  }
  
  # This snippet is needed because some of the ticker symbols include are invalid variable names
  dfSyms$Symbol <- str_replace_all(dfSyms$Symbol, "[^[:alnum:]]", "")

  xtsData <- get(dfSyms$Symbol[1])
  for(idx in 2:nrow(dfSyms)){
    xtsData <- merge(xtsData,  get(dfSyms$Symbol[idx]) )
  }
  
  dfData <- data.frame(xtsData)
  
  # These two lines evenly space all the data at a daily interval
  zooData <- zoo(dfData, as.Date(rownames(dfData)))
  zooData <- merge.zoo(zooData, zoo(,seq(start(zooData), end(zooData), by=1)), all=TRUE)
  dfData <- data.frame(zooData)

  # Save it off
  save(dfData,file="dfData.Rda")
  
}else{
  load("dfData.Rda")
}

```


## Interpolate data

The data is interpolated to a daily basis. This is done because I also incorporate other data sampled at different intervals. When performing correlations, it is easier if all the data has the same time basis.

```{r tidydata}

# Linearly interpolate the data, add normalized columns (this helps speed portfolio calcs)

for (col_name in names(dfData))
{
    if (is.numeric(dfData[,col_name]))
    {
      dfData[,col_name] <- na.approx(dfData[,col_name], rule=2)
      
    }
}

```


## Add date column

```{r adddate}
# This speeds some of the upcoming operations
dfData$date = as.Date(rownames(dfData))

```



## Normalize data

```{r normdata}

# Linearly interpolate the data, add normalized columns (this helps speed portfolio calcs)

for (col_name in names(dfData))
{
    if (is.numeric(dfData[,col_name]))
    {
      # Normalize data
      strColNorm <- paste(col_name, "_Norm", sep="") 
      dfData[, strColNorm] <- dfData[,col_name] / dfData[dfData$date==as.Date(strDateStart),col_name]

    }
}

```

# Calculate the returns

## Calculate the time based returns on annual basis

In this analysis all the returns are calculated on an annual basis. It means there is a year of dead data, but it eliminates seasonal variation in data (think: Christmas).

```{r Returncalcs}
for (strName in names(dfData)){
  
  if( is.numeric(dfData[, strName])){
    
    # Year over year returns
    strNameYoY <- paste(strName,"_YoY",sep="")
    dfData[strNameYoY] <- CalcYoY(dfData, strName, iRetPd)

    # log diff returns
    strNameLog <- paste(strName, "_LogRet", sep="")
    dfData[strNameLog] <- CalcLogRet(dfData, strName, iRetPd)
    
    strRootSym <- strName
    strSuffix <- ""
    if( grepl("\\.", strName)){
      strRootSym <-  substr(strName, 1, regexpr("\\.", strName)-1)
      strSuffix <- substr(strName, regexpr("\\.", strName)+1, nchar(strName))
      strSuffix <- paste(" (", strSuffix, ")", sep="")
    }
    
    strNewDesc <- dfSyms[grep(paste("^",strRootSym,"$",sep=""), dfSyms$Symbol),]$Desc
    dfSyms <- rbind(dfSyms, data.frame(Symbol=strNameYoY, Source="Calc", 
                Desc=paste(strNewDesc, strSuffix, " Year over Year", sep=""), yLabel="Percent" ))
    
    
      
  }

}

```


## Truncate data

A few last calculations to truncate and aggregrate the data

```{r portUpdate}

# Truncate by date, if needed
dfData <- with(dfData, dfData[date>=strDateStart,])


```


## Summarize returns

In constructing the portfolio it is importance to understand the blend of funds. We want maximum returns for minimum volatility. That's not possible in a perfect world so we plot the returns and volatility to find a nice balance in the portfolio.

```{r Returns}

# Linearly interpolate the data, add normalized columns (this helps speed portfolio calcs)

for (col_name in names(dfData))
{
    if (is.numeric(dfData[,col_name]))
    {

      # Split the name ("USGFG.Close"" is "USGFG"" and "Close"")
      lstSyms <- lstSymSplit(col_name)
        
      # Only if there is two terms
      if( length(lstSyms) > 1){

        if( lstSyms[2] == 'Close_YoY' ){

          # Return is the mean of the series
          dfSyms[dfSyms$Symbol==lstSyms[1],'ExpReturn'] <- mean(dfData[, col_name])

          # Volatility is the standard deviation of the series.
          dfSyms[dfSyms$Symbol==lstSyms[1],'Volatility'] <- sd(dfData[, col_name])
          
        }
      }

    }
}



```

# Define the various portfolios

This code is evaluating two portfolios. "pf_USAA_Legacy" is the original portfolio and "pf_USAA_New" is the one I'm trying out.

```{r defPorts}

dfSyms$pf_USAA_Legacy <- 0
dfSyms[dfSyms$Symbol=='USSPX',]$pf_USAA_Legacy <-  0.4948
dfSyms[dfSyms$Symbol=='USAWX',]$pf_USAA_Legacy <-  0.1429
dfSyms[dfSyms$Symbol=='USISX',]$pf_USAA_Legacy <-  0.2383
dfSyms[dfSyms$Symbol=='UFSGX',]$pf_USAA_Legacy <- 0.1240

dfSyms$pf_USAA_New <- 0
dfSyms[dfSyms$Symbol=='USPRX',]$pf_USAA_New <-  0.00000
dfSyms[dfSyms$Symbol=='USSPX',]$pf_USAA_New <-  0.4500
dfSyms[dfSyms$Symbol=='USAWX',]$pf_USAA_New <-  0.0000
dfSyms[dfSyms$Symbol=='USISX',]$pf_USAA_New <-  0.00000
dfSyms[dfSyms$Symbol=='USNQX',]$pf_USAA_New <- 0.5500
dfSyms[dfSyms$Symbol=='USCGX',]$pf_USAA_New <- 0.0000
dfSyms[dfSyms$Symbol=='UFSGX',]$pf_USAA_New <- 0.0000
dfSyms[dfSyms$Symbol=='USCCX',]$pf_USAA_New <- 0.0001
dfSyms[dfSyms$Symbol=='USAAX',]$pf_USAA_New <- 0.0000

dfSyms$pf_Roth_Legacy <- 0
dfSyms[dfSyms$Symbol=='VTWO',]$pf_Roth_Legacy <-  0.0947
dfSyms[dfSyms$Symbol=='VFINX',]$pf_Roth_Legacy <-  0.5051
dfSyms[dfSyms$Symbol=='TMFGX',]$pf_Roth_Legacy <-  0.2568
dfSyms[dfSyms$Symbol=='IWM',]$pf_Roth_Legacy <-  0.0000
dfSyms[dfSyms$Symbol=='QQQ',]$pf_Roth_Legacy <-  0.0000
dfSyms[dfSyms$Symbol=='HAINX',]$pf_Roth_Legacy <-  0.0000
dfSyms[dfSyms$Symbol=='VEU',]$pf_Roth_Legacy <-  0.0000

dfSyms$pf_Roth_New <- 0
dfSyms[dfSyms$Symbol=='VTWO',]$pf_Roth_New <-   0.0000
dfSyms[dfSyms$Symbol=='VFINX',]$pf_Roth_New <-  0.3000
dfSyms[dfSyms$Symbol=='TMFGX',]$pf_Roth_New <-  0.0500
dfSyms[dfSyms$Symbol=='IWM',]$pf_Roth_New <-    0.0000
dfSyms[dfSyms$Symbol=='QQQ',]$pf_Roth_New <-    0.4500
dfSyms[dfSyms$Symbol=='HAINX',]$pf_Roth_New <-  0.0000
dfSyms[dfSyms$Symbol=='CAMIX',]$pf_Roth_New <-  0.0000
dfSyms[dfSyms$Symbol=='IVOO',]$pf_Roth_New <-   0.1500
dfSyms[dfSyms$Symbol=='VEU',]$pf_Roth_New <-    0.0000
dfSyms[dfSyms$Symbol=='SPY',]$pf_Roth_New <-    0.0000
dfSyms[dfSyms$Symbol=='VIGI',]$pf_Roth_New <-   0.0000
dfSyms[dfSyms$Symbol=='SCHD',]$pf_Roth_New <-   0.0000
dfSyms[dfSyms$Symbol=='VYM',]$pf_Roth_New <-    0.0500

dfSyms$pf_Div <- 0
dfSyms[dfSyms$Symbol=='VFINX',]$pf_Div <-  0.1000
dfSyms[dfSyms$Symbol=='SCHD',]$pf_Div <-  0.1000
dfSyms[dfSyms$Symbol=='VYM',]$pf_Div <-  0.1000
dfSyms[dfSyms$Symbol=='NOBL',]$pf_Div <-  0.1000


# This part normalizes all the data, in case there is a typo or 
# numerical rounding that results in somthing less than 100%
for (col_name in names(dfSyms)){
  
  if( length(grep('pf_', col_name))){
    dfSyms[,col_name] <- dfSyms[,col_name] / sum(dfSyms[,col_name])
  }
  
}

# Update the returns
dfData <- pfUpdateReturn("pf_USAA_Legacy", dfData, dfSyms)
dfData <- pfUpdateReturn("pf_USAA_New", dfData, dfSyms)
dfData <- pfUpdateReturn("pf_Roth_Legacy", dfData, dfSyms)
dfData <- pfUpdateReturn("pf_Roth_New", dfData, dfSyms)
dfData <- pfUpdateReturn("pf_Div", dfData, dfSyms)

kable(dfSyms[(dfSyms$pf_USAA_Legacy>0.0) | (dfSyms$pf_USAA_New>0.0)
             | (dfSyms$pf_Roth_New>0.0)| (dfSyms$pf_Div>0.0),])

```

# Legacy USAA Portfolio

## Time series analysis

This code plots out the legacy portfolio performance. It is not great, lagging behind the S&P 500. Ideally it would keep up with or beat that bench mark.

```{r portPlot1_Time}

datay <- "pf_USAA_Legacy"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

  
```

## Risk return analysis

This section plots the funds on return and volatility axis. An ideal fund would be in the upper left corner of the plot. The USISX is in the wrong place, low returns and high risk. Uggh.

```{r portPlot1_RiskReturn}

rrData <- plotRR(datay)
rrData[[2]]

```

The same data as in the plot, but presented as a table.

```{r port1_RRTable}

tblPort(rrData, datay)

```

## Time series analysis (Optimized)

As a final check, plot the optimized portfolio against the inputted portfolio and the benchmark.

```{r portPlot1_Check}

datayOpt <- paste(datay, "_Opt", sep="")
dfData <- pfUpdateReturn(datayOpt, dfData, rrData[[1]])
ylim <- c(0, 4)
myPlot <- plotSingleBench(datay, ylim)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datayOpt, colour=shQuote(datayOpt)), na.rm = TRUE)

```

## Correlation analysis

The more funds are negatively correlated, the more volatility can be reduced. This plot shows how the funds in the portfolio correlate.

```{r corplot1, echo=TRUE, fig.width=6,fig.height=6.5}

plotCorr(datay)

```

# New USAA Portfolio

## Time series analysis

This one is on a much better track, actually beating the S&P 500 index. Looking at the plots below, I think the mix should be 45% USSPX and 55% USNQX. This should have less volatility and better returns than the existing portfolio, although the models suggest that even less volatility could be had by including USCCX. I have hesitation about the bonds, with all the upcoming rate hikes by the Federal Reserve. But USCCX is highly uncorrelated...need to revisit this in the fall of 2017, after the moves by the Fed.

```{r portPlot2}

datay <- "pf_USAA_New"
ylim <- c(0, 4)
plotSingleBench(datay, ylim)

```

## Risk return analysis

This section plots the funds on return and volatility axis. An ideal fund would be in the upper left corner of the plot.

```{r portPlot2_RiskReturn}

datay <- "pf_USAA_New"
rrData <- plotRR(datay)
rrData[[2]]

```

This table shows the portfolio. Column `r datay` shows the inputted portfolio and `r paste(datay,"_Opt", sep="")` shows the portfolio with the same returns, but less risk. 

```{r port2_RRTable}

tblPort(rrData, datay)

```

## Time series analysis (Optimized)

As a final check, plot the optimized portfolio against the inputted portfolio and the benchmark.


```{r portPlot2_Check}

datay <- "pf_USAA_New"
datayOpt <- paste(datay, "_Opt", sep="")
dfData <- pfUpdateReturn(datayOpt, dfData, rrData[[1]])
ylim <- c(0, 4)
myPlot <- plotSingleBench(datay, ylim)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datayOpt, colour=shQuote(datayOpt)), na.rm = TRUE)

```

## Correlation analysis

The more funds are negatively correlated, the more volatility can be reduced. This plot shows how the funds in the portfolio correlate.

```{r corplot2, echo=TRUE, fig.width=6,fig.height=6.5}

plotCorr(datay)

```



# Legacy Roth Portfolio

## Time series analysis

Uggh, just uggh. This one is performing much worse than I expected.

```{r portPlotRoth1}

datay <- "pf_Roth_Legacy"
ylim <- c(0, 4)
plotSingleBench(datay, ylim)

```

## Risk return analysis

This section plots the funds on return and volatility axis. An ideal fund would be in the upper left corner of the plot.

```{r portPlot3_RiskReturn}

rrData <- plotRR(datay)
rrData[[2]]

```

This table shows the portfolio. Column `r datay` shows the inputted portfolio and `r paste(datay,"_Opt", sep="")` shows the portfolio with the same returns, but less risk. 

```{r port3_RRTable}

tblPort(rrData, datay)

```

## Time series analysis (Optimized)

As a final check, plot the optimized portfolio against the inputted portfolio and the benchmark.


```{r portPlot3_Check}

datayOpt <- paste(datay, "_Opt", sep="")
dfData <- pfUpdateReturn(datayOpt, dfData, rrData[[1]])
ylim <- c(0, 4)
myPlot <- plotSingleBench(datay, ylim)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datayOpt, colour=shQuote(datayOpt)), na.rm = TRUE)

```

## Correlation analysis

The more funds are negatively correlated, the more volatility can be reduced. This plot shows how the funds in the portfolio correlate.

```{r corplot3, echo=TRUE, fig.width=6,fig.height=6.5}

plotCorr(datay)

```

# New Roth Portfolio

## Time series analysis

CAPM portfolio theory suggests that only TMFGX, VTWO, and QQQ are needed, but these haven't performed quite as well as the mix below. Looking at the correlation plot only TMFGX does not correlate with the other funds so it is a must have. The final mix will be 30% VFINX, 5% of TMFGX (low volatility so like a bond fund), 5% of VYM (dividend exposure), 15% to IVOO (mid cap), and 45% to QQQ.

```{r portPlotRothNew}

datay <- "pf_Roth_New"
ylim <- c(0, 4)
plotSingleBench(datay, ylim)

```

## Risk return analysis

This section plots the funds on return and volatility axis. An ideal fund would be in the upper left corner of the plot.

```{r portPlot4_RiskReturn}

rrData <- plotRR(datay)
rrData[[2]]

```

This table shows the portfolio. Column `r datay` shows the inputted portfolio and `r paste(datay,"_Opt", sep="")` shows the portfolio with the same returns, but less risk. 

```{r port4_RRTable}

tblPort(rrData, datay)

```

## Time series analysis (Optimized)

As a final check, plot the optimized portfolio against the inputted portfolio and the benchmark.


```{r portPlot4_Check}

datayOpt <- paste(datay, "_Opt", sep="")
dfData <- pfUpdateReturn(datayOpt, dfData, rrData[[1]])
ylim <- c(0, 4)
myPlot <- plotSingleBench(datay, ylim)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datayOpt, colour=shQuote(datayOpt)), na.rm = TRUE)

```

## Correlation analysis

The more funds are negatively correlated, the more volatility can be reduced. This plot shows how the funds in the portfolio correlate.

```{r corplot4, echo=TRUE, fig.width=6,fig.height=6.5}

plotCorr(datay)

```


# Dividend Portfolio

## Time series analysis

This portfolio looks at a weighting favoring dividend funds..

```{r portPlotDiv}

datay <- "pf_Div"
ylim <- c(0, 4)
plotSingleBench(datay, ylim)

```

## Risk return analysis

This section plots the funds on return and volatility axis. An ideal fund would be in the upper left corner of the plot.

```{r portPlot5_RiskReturn}

rrData <- plotRR(datay, TRUE)
rrData[[2]]

```

This table shows the portfolio. Column `r datay` shows the inputted portfolio and `r paste(datay,"_Opt", sep="")` shows the portfolio with the same returns, but less risk. 

```{r port5_RRTable}

tblPort(rrData, datay)

```

## Time series analysis (Optimized)

As a final check, plot the optimized portfolio against the inputted portfolio and the benchmark.


```{r portPlot5_Check}

datayOpt <- paste(datay, "_Opt", sep="")
dfData <- pfUpdateReturn(datayOpt, dfData, rrData[[1]])
ylim <- c(0, 4)
myPlot <- plotSingleBench(datay, ylim)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datayOpt, colour=shQuote(datayOpt)), na.rm = TRUE)

```

## Correlation analysis

The more funds are negatively correlated, the more volatility can be reduced. This plot shows how the funds in the portfolio correlate.

```{r corplot5, echo=TRUE, fig.width=6,fig.height=6.5}

plotCorr(datay)

```

# Appendix


## USSPX, USAA S&P 500 Index Member



```{r USPRX}

datay <- "USSPX.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```

## USSPX Year over year, USAA S&P 500 Index Member


```{r USPRX_YoY}

datay <- "USSPX.Close_Norm_YoY"
ylim <- c(-1, 1)
plotSingleQuick(datay, ylim)

```

## USISX, USAA Income Stock


```{r USISX}

datay <- "USISX.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```

## USISX Year over year, USAA Income Stock


```{r USISX_YoY}

datay <- "USISX.Close_Norm_YoY"
ylim <- c(-1, 1)
plotSingleQuick(datay, ylim)

```

## UFSGX, USAA First Start Growth

```{r UFSGX}

datay <- "UFSGX.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```

## UFSGX Returns, USAA First Start Growth


```{r UFSGX_Return}

datay <- "UFSGX.Close_Norm_YoY"
ylim <- c(-1, 1)
plotSingleQuick(datay, ylim)

```

## USCCX, USAA Cornerstone Conservative

```{r USCCX}

datay <- "USCCX.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```

## USCCX Year over year, USAA Cornerstone Conservative

```{r USCCX_YoY}

datay <- "USCCX.Close_Norm_YoY"
ylim <- c(-1, 1)
plotSingleQuick(datay, ylim)

```

## VTWO, Vanguard Russell 2000 ETF

Not as much history on this one as I would like.

```{r VTWO}

datay <- "VTWO.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```



## IWM, iShares Russell 2000

Not as much history on this one as I would like.

```{r IWM}

datay <- "IWM.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```

## SPY, SPDR S&P 500 ETF Trust

Should be real close to the S&P 500

```{r SPY}

datay <- "SPY.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```

## VFINX, Vanguard 500 Index Investor

Should be real close to the S&P 500

```{r VFINX}

datay <- "VFINX.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)
```
## SCHD, Schwab U.S. Dividend Equity ETF

Nice little pick of divident paying stocks on Dow Jones Dividend 100 index

```{r SCHD}

datay <- "SCHD.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)
```

```{r SCHD_YoY}

datay <- "SCHD.Close_Norm_YoY"
ylim <- c(-1, 1)
plotSingleQuick(datay, ylim)

```

## IVOO, Vanguard S&P Mid-Cap 400 ETF 

The investment seeks to track the performance of a benchmark index that measures the investment return of mid-capitalization stocks in the United States. The fund employs an indexing investment approach designed to track the performance of the S&P MidCap 400® Index. The index measures the performance of mid-capitalization stocks in the United States. The Advisor attempts to replicate the target index by investing all, or substantially all, of its assets in the stocks that make up the index, holding each stock in approximately the same proportion as its weighting in the index.

Expense Ratio: 0.15%

```{r IVOO}

datay <- "IVOO.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)
```
```{r IVOO_YoY}

datay <- "IVOO.Close_Norm_YoY"
ylim <- c(-1, 1)
plotSingleQuick(datay, ylim)

```

## VYM, Vanguard High Dividend Yield ETF

Another nice little pick of divident paying stocks on Dow Jones Dividend 100 index. Expense ratio: 0.08%

```{r VYM}

datay <- "VYM.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)
```

```{r VYM_YoY}

datay <- "VYM.Close_Norm_YoY"
ylim <- c(-1, 1)
plotSingleQuick(datay, ylim)

```

## CAMIX, Cambiar International Equity Inv

Some exposure to foriegn stocks

```{r CAMIX}

datay <- "CAMIX.Close_Norm"
ylim <- c(0, 3)
plotSingleBench(datay, ylim)

```


