---
title: "RecessionModels"
author: "Brian Howard"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    smart: false
output_file: "RecessionModels.html"    
---

```{r setup, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options("getSymbols.warning4.0"=FALSE)

```

```{r libraries, echo=FALSE, message=FALSE}

library(tidyverse)
library(tidyquant)
library(Hmisc)
library(gridExtra)
library(signal)
library(caret)
library(data.table)
library(lubridate)
#install.packages("NeuralNetTools")
library(NeuralNetTools)

```

```{r helper functions, echo=FALSE}

# Call helper functions
source("plotHelper.r")
source("modelHelper.r")

```


```{r load.data, echo=FALSE}

str.data.dir <- "C:/Users/Rainy/OneDrive/Documents/IssaquahDynamical/Datasets/2529RS0082_HistEconData"
out.file <- file.path(str.data.dir, "RecessionIndicator_Buffer.RData")

# Load data
load(out.file)

rm(str.data.dir)
rm(out.file)

```


```{r plot.limits, echo=FALSE}

# Define the plotting ranges
dt.recent = as.Date("2017-01-01")

# Define the start of the prediction
dt.start.prediction <- as.Date("01/02/1962", "%d/%m/%Y")

# Back test range
dt_start_back_test = as.Date('1960-01-01')

# Define the S&P 500 limits
source("SetGSPCMax.r")

```

```{r clean_data, echo=FALSE}

# Keep only the symbols needed by this code
df.symbols <- dplyr::filter(df.symbols,
                            string.symbol %in% c("USREC",
                                                 "GDP",
                                                 "GDPC1",
                                                 "CPROFIT",
                                                 "NPPTTL",
                                                 "ADPWNUSNERSA",
                                                 "CCSA",
                                                 "PCE",
                                                 "U6RATE",
                                                 "UNRATE",
                                                 "ICSA",
                                                 "POPTHM",
                                                 "CLF16OV"))

if( exists("b.refresh.models")){
  rm(b.refresh.models)
}

if( exists("bRefresh")){
  rm(bRefresh)
}

if( exists("d.income")){
  rm(d.income)
}

if( exists("d.Russell.max")){
  rm(d.Russell.max)
}

if( exists("dt.date")){
  rm(dt.date)
}

if( exists("dt.date.c")){
  rm(dt.date.c)
}

if( exists("i.cols")){
  rm(i.cols)
}

if( exists("idx")){
  rm(idx)
}

if( exists("iDays")){
  rm(iDays)
}

if( exists("list.margin.names")){
  rm(list.margin.names)
}

if( exists("list.options.names")){
  rm(list.options.names)
}

if( exists("str.name")){
  rm(str.name)
}

if( exists("str.safe.name")){
  rm(str.safe.name)
}

if( exists("str_symbol")){
  rm(str_symbol)
}

if( exists("str.symbol")){
  rm(str.symbol)
}

if( exists("str_out")){
  rm(str_out)
}

if( exists("str.symbol.raw")){
  rm(str.symbol.raw)
}

if( exists("str_symbol.raw")){
  rm(str_symbol.raw)
}

if( exists("ylimBackTest")){
  rm(ylimBackTest)
}

if( exists("df.farm.income")){
  rm(df.farm.income)
}

```

## Feature Extraction

With the raw data downloaded, some of the interesting features can be extracted.
The first step is reconcile the time intervals. Some of the data is released
monthly and some daily. I chose to interpolate all data to a daily interval. The
first section of code adds the daily rows to the data frame.

The code performs interpolation for continuous data or carries it forward for
binary data like the recession indicators.

```{r aggsyms}

source("calcInterpolate.r")
df.data <- calcInterpolate(df.symbols)

```

## Truncate data

```{r data.trunc, echo=FALSE, fig.width = 10, fig.asp = .62}

# A very few data series do go back to 1854, but most
# don't even go past WWII so truncate the data frame.
df.data <- df.data[as.Date(rownames(df.data)) > as.Date("1940-01-01"),]

```

## Create aggregate series

Some analysis requires that two or more series be combined. For example, normallizing debt by GDP to get a sense of the proportion of debt to the total economy helps understand the debt cycle.

```{r create aggregate, echo=FALSE}

source("calcAggregateSeries.r")

```

Year over year, smoothed derivative, and log trends tend to smooth out seasonal variation. It gets used so often that I do this for every series downloaded.

```{r calcsYoYSmoothLog}

source("calcFeatures.r")
lst.df <- calcFeatures(df.data, df.symbols)
df.data <- lst.df[[1]]
df.symbols <- lst.df[[2]]

```


```{r calc.features.for.aggregate, echo=FALSE}

# Calculate the features for the aggregated series
source("calcFeaturesAggregate.r")

```


# Recession initiation and termination dates/times

Build the recession and recession initiation dates

```{r recframe}

source("calcRecession.r")

```

# Recession models

## Input data series

The labor market is one of the clearest, most reliable signals of broad economic
turning points—but it lags.

```{r rec_model_unrate_base, echo=FALSE, fig.width = 10, fig.asp = .62}

# Define the symbols in the plot
lst_syms <- c("UNRATE.Value", "UNRATE.Value__Smooth")

if (require_columns(df.data, lst_syms)) {
  ylim <- c(2, 12)
  b.legend <- TRUE
  b.percentile <- TRUE
  myPlot <- plotSingle(
    dfRecession,
    df.data,
    "date",
    datay = lst_syms[[1]],
    getPlotTitle(df.symbols, lst_syms[[1]]),
    "Date",
    getPlotYLabel(df.symbols, lst_syms[[1]]),
    c(as.Date("1jan1950", "%d%b%Y"), Sys.Date()),
    ylim,
    b.legend,
    b.percentile
  )
  myPlot + geom_line(
    data = df.data,
    aes(
      x = .data[["date"]],
      y = .data[[lst_syms[[2]]]],
      colour = getPlotYLabel(df.symbols, lst_syms[[2]])
    ),
    na.rm = TRUE
  )
}

```

```{r rec_model_GDP_base, echo=FALSE, fig.width = 10, fig.asp = .62}

# Define the symbols in the plot
lst_syms <- c("GDP.Value")

# Do they exist in the data frame?
if ( require_columns(df.data, lst_syms ) ){

  my_plot <-
    plotSingle(
      dfRecession,
      df.data,
      "date",
      datay = lst_syms[[1]],
      getPlotTitle(df.symbols, lst_syms[[1]]),
      "Date",
      getPlotYLabel(df.symbols, lst_syms[[1]]),
      c(dt.start.prediction, Sys.Date()),
      ylim = c(1, round_up_to(max(df.data[[lst_syms[[1]]]]),5000)),
      b.legend = TRUE,
      b.percentile = FALSE,
      b.long.legend = TRUE
    )

  print(my_plot)
  
  # Free up memory
  if( exists("my_plot")){
    rm(my_plot)
  }
}


# Free up memory
if( exists("lst_syms")){
  rm(lst_syms)
}


```

```{r rec_model_cprofit_by_gdp_base, echo=FALSE, fig.width = 10, fig.asp = .62}

# Define the symbols in the plot
lst_syms <- c("CPROFIT__by__GDP")

# Do they exist in the data frame?
if ( require_columns(df.data, lst_syms ) ){

  my_plot <-
    plotSingle(
      dfRecession,
      df.data,
      "date",
      datay = lst_syms[[1]],
      getPlotTitle(df.symbols, lst_syms[[1]]),
      "Date",
      getPlotYLabel(df.symbols, lst_syms[[1]]),
      c(dt.start.prediction, Sys.Date()),
      ylim = c(5, round_up_to(max(df.data[[lst_syms[[1]]]]),5)),
      b.legend = TRUE,
      b.percentile = FALSE,
      b.long.legend = TRUE
    )

  print(my_plot)
  
  # Free up memory
  if( exists("my_plot")){
    rm(my_plot)
  }
}


# Free up memory
if( exists("lst_syms")){
  rm(lst_syms)
}


```



