---
title: "Recession Indicators"
author: "Brian Howard"
date: "Feb 26, 2018"
output: 
  html_document:
    toc: true
    toc_depth: 2
    smart: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("getSymbols.warning4.0"=FALSE)
```

```{r libraries, echo=FALSE, message=FALSE}
library(UsingR)
library(quantmod)
library(ggplot2)
library(grid)
library(gridExtra)
library(knitr)
library(tools)
library(zoo)
library(signal)
library(stringr)
library(corrplot)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(Quandl)
# library(devtools)
# install_github("andrewuhl/RollingWindow")
library(RollingWindow)
library(gtable)
```

This function provides a common format for the plots

```{r plothelp}
# ------------------------------------------------------------------------------
# Define the generic single plane plot function
plotSingle <- function(datadf_rec, datadf, datax, datay, titlelabel, xlabel, ylabel, xlim, ylim, bLegend){
  
  myplot <- ggplot()+
    theme(plot.background = element_rect(fill = 'white', colour = 'white')) +
    theme(panel.background = element_rect(fill = 'white', colour = 'grey')) +
    theme(panel.grid.major.x = element_blank()) +
    theme(panel.grid.major.y = element_line(colour="grey", size=0.5)) +
    geom_line(data=datadf, aes_string(x=datax, y=datay, colour = factor(datay)), 
              na.rm = TRUE, size = 0.7) +
    scale_colour_manual(values = c("black","blue", "green", "red", "magenta")) +
    guides(colour = guide_legend("Series"), size = guide_legend("Series"), shape = guide_legend("Series")) +
    geom_rect(data=datadf_rec,
              aes(xmin=start,xmax=end, ymin=-Inf, ymax=Inf),
              fill="grey", alpha=0.3, na.rm = TRUE) +
    scale_fill_continuous(name = "V") +
    #geom_smooth(method = "lm") + 
    ggtitle(titlelabel) +
    labs(x=xlabel, y = ylabel) +
    scale_x_date(limits = xlim ) + 
    scale_y_continuous(limits = ylim) +
    if( bLegend){
      theme(legend.position = "top")
    }else{
      theme(legend.position = "none")
    }
  

  return(myplot)
}

# ------------------------------------------------------------------------------
# Define the function for calculating year over year growth. 
CalcYoY <- function (datadf, strCol, iPeriods){
  Nrow <- nrow(datadf)
  GrowthRateYoY <- rep(0,Nrow)
  GrowthRateYoY[(iPeriods+1):Nrow] <- diff(as.matrix(datadf[[strCol]]), lag = iPeriods)
  GrowthRateYoY <- (GrowthRateYoY / datadf[[strCol]])*100
  return(GrowthRateYoY)
}


# Small helper function to get the symbol description
getPlotTitle <- function(datay){
  strTitle <-  paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$Desc)
  return(strTitle)
}

getPlotYLabel <- function(datay){
  strY <- dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$yLabel
  return(strY)
}

plotSingleQuickRecent <- function(datay, ylim){
  
  plotSingle(dfRecession, dfData, "date", datay, 
           paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$Desc), "Date", 
           dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, c(as.Date("1jan2000","%d%b%Y"), Sys.Date()), ylim, FALSE)
  
}

plotSingleQuickRecentRecent <- function(datay, ylim){
  
  plotSingle(dfRecession, dfData, "date", datay, 
           paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$Desc), "Date", 
           dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, c(as.Date("1jan2010","%d%b%Y"), Sys.Date()), ylim, FALSE)
  
}

plotSingleQuickModern <- function(datay, ylim){
  
  plotSingle(dfRecession, dfData, "date", datay, 
           paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$Desc), "Date", 
           dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, c(as.Date("1jan1960","%d%b%Y"), Sys.Date()), ylim, FALSE)
  
}

plotSingleQuick <- function(datay, ylim){
  
  plotSingle(dfRecession, dfData, "date", datay, 
           paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""),
           dfSyms$Symbol),]$Desc), "Date", 
           dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, FALSE)
  
}

plotSingleQuickDate <- function(datay, ylim, dtStart){
  
  plotSingle(dfRecession, dfData, "date", datay, 
           paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$Desc), "Date", 
           dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, c(dtStart, Sys.Date()), ylim, FALSE)
  
}


plotSingleBackPred <- function(datay, ylim, dfPred){
  
  plotSingle(dfRecession, dfData, "date", datay, 
           paste(datay, " | ", dfSyms[grep(paste("^", datay, "$", sep=""), dfSyms$Symbol),]$Desc), "Date", 
           dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, FALSE) +
  geom_rect(data=dfPred,  aes(xmin=predStart, xmax=predEnd, ymin=-Inf, ymax=Inf),
                fill="red", alpha=0.2, na.rm = TRUE)
  
}

plotSingleBackGrowth <- function(datay, ylim){
  
  myPlot <- plotSingle(dfRecession, dfData, "date", datay, 
              paste(datay, " | ", dfSyms[grep(datay, dfSyms$Symbol),]$Desc), "Date", 
              dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, TRUE) +   
            geom_line(data=dfData, aes_string(x="date", y="eqBase", colour=shQuote("eqBase")), na.rm = TRUE)
  
  return(myPlot)

}

# Define the generic single plane plot function
plotXvY <- function(dfData, datax, datay, titlelabel, xlabel, ylabel, xlim, ylim, bLegend, bFitLinear, dtStart){
  
  # Select data of interest
  dfPlot <- dfData[dfData$date>=dtStart,]
  
  # Include date range
  titlelabel <- paste(titlelabel, "\nDate range: ", dtStart, " to ", tail(dfPlot$date, n=1), sep="")
  
  # Include a linear fit?
  if( bFitLinear ){
    formFit <- as.formula(paste(datay, '~', datax))
    lm_fit <- lm(formFit, data = dfPlot)
    titlelabel <- paste(titlelabel, "\nAdj R2=", signif(summary(lm_fit)$adj.r.squared, 5), 
                        " | Intercept =",signif(lm_fit$coef[[1]],5 ), 
                        " | Slope =",signif(lm_fit$coef[[2]], 5), sep="")
  }
  
  myplot <- ggplot()+
    theme(plot.background = element_rect(fill = 'white', colour = 'white')) +
    theme(panel.background = element_rect(fill = 'white', colour = 'grey')) +
    theme(panel.grid.major.x = element_line(colour="grey", size=0.5)) +
    theme(panel.grid.major.y = element_line(colour="grey", size=0.5)) +
    geom_point(data=dfPlot, aes_string(x=datax, y=datay, colour = factor(datay)), 
              na.rm = TRUE, size = 0.7) +
    stat_smooth(data=dfPlot, aes_string(x=datax, y=datay, colour = factor(datay)), method = "lm", col = "red") +
    scale_colour_manual(values = c("black","blue", "green")) +
    guides(colour = guide_legend("Series"), size = guide_legend("Series"), shape = guide_legend("Series")) +
    scale_fill_continuous(name = "V") +
    ggtitle(titlelabel) +
    labs(x=xlabel, y = ylabel) +
    scale_x_continuous(limits = xlim ) +
    scale_y_continuous(limits = ylim) +
    if( bLegend ){
      theme(legend.position = "top")
    }else{
      theme(legend.position = "none")
    }

  return(myplot)
}

```

```{r plotbackhelp, echo=FALSE}

plotBack <- function (dataTrade, dataRet, dataEq, dfPred){

  
  ylim <- c(0, 1)
  p1 <- plotSingleBackPred(dataTrade, ylim, dfPred)

  ylim <- c(-0.1, 0.1)
  p2 <- plotSingleQuick(dataRet, ylim)
  

  ylim <- c(0, 1000)
  p3 <- plotSingleBackGrowth (dataEq, ylim)
  
  dPercent <- 100*( tail(dfData$eqRec,1)-tail(dfData$eqBase,1))/tail(dfData$eqBase,1) 
  
  grid.arrange(p1, p2, p3, ncol = 1, top = paste("Recession Initiation Rule | Baseline Growth = ", 
                                                 sprintf('%0.2f', tail(dfData$eqBase,1)), 
                                                 " | Rule Growth = ",  
                                                 sprintf('%0.2f', tail(dfData$eqRec,1)), 
                                                 "\nPercent Improvement: ",
                                                 sprintf('%0.1f%s', dPercent, "%") , sep="") )

}

```

# Introduction

Wouldn't it be fun to see into future? I think so. For one of my machine learning classes we had a project that consumed financial data. I have extended that project to use machine learning to see if an indicator, or predictor, can be found that identifies market tops that occur prior to recessions. Then I use the model to build a trading strategy and backtest it to see how it performs.

# Get Economic and Financial Data

Data is pulled from several sources include FRED, yahoo, and Google. The code below shows an example that pulls in the consumer price index (CPI) from the FRED. 

```{r defsyms, echo=FALSE}
Symbol = c("CPIAUCSL", "USREC")
Source = c("FRED", "FRED")
Desc = c("Consumer Price Index for All Urban Consumers: All Items",
         "NBER based Recession Indicators")
yLabel = c("Index 1982-1984=100","+1 or 0")
dfSyms = data.frame(Symbol, Source, Desc, yLabel, stringsAsFactors = FALSE)

dfSyms <- rbind(dfSyms, data.frame(Symbol="UNRATE", Source="FRED", 
              Desc="Civilian Unemployment Rate U-3", 
              yLabel="Percent" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="U6RATE", Source="FRED", 
              Desc="Total unemployed + margin + part-time U-6", 
              yLabel="Percent" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="HNONWPDPI", Source="FRED", 
              Desc="Household Net Worth, percent Dispsable Income", 
              yLabel="Percent" ))

dfSyms <- rbind(dfSyms, data.frame(Symbol="INDPRO", Source="FRED", 
              Desc="Industrial Production Index", 
              yLabel ="Index 2012=100" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="RRSFS", Source="FRED", 
              Desc="Real Retail and Food Services Sales", 
              yLabel ="Millions of Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="RSALES", Source="FRED", 
              Desc="Real Retail Sales (DISCONTINUED)",
              yLabel ="Millions of Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="W875RX1A020NBEA", Source="FRED", 
              Desc="Real personal income\nexcluding current transfer receipts",
              yLabel = "Billions of Chained 2009 Dollars") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="RPI", Source="FRED", 
              Desc="Real personal income",
              yLabel = "Billions of Chained 2009 Dollars") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="PCOPPUSDM", Source="FRED", 
              Desc="Global price of Copper",
              yLabel = "U.S. Dollars per Metric Ton") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="NOBL", Source="yahoo", 
              Desc="ProShares S&P 500 Dividend Aristocrats (NOBL)",
              yLabel = "BATS Real Time Price") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="BUSLOANS", Source="FRED", 
              Desc="Commercial and Industrial Loans,\n All Commercial Banks",
              yLabel = "Billions of U.S. Dollars") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="NCBDBIQ027S", Source="FRED", 
              Desc="Nonfinancial corporate\n business; debt securities; liability, Level",
              yLabel = "Millions of Dollars") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="REALLNNSA", Source="FRED", 
              Desc="Real Estate Loans, All Commercial Banks",
              yLabel = "Billions of U.S. Dollars") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="CONSUMERNSA", Source="FRED", 
              Desc="Consumer Loans, All Commercial Banks",
              yLabel = "Billions of U.S. Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="DGS10", Source="FRED", 
              Desc="10-Year Treasury Constant Maturity Rate",
              yLabel = "Percent" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="DGS1", Source="FRED", 
              Desc="1-Year Treasury Constant Maturity Rate",
              yLabel = "Percent" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="TB3MS", Source="FRED", 
              Desc="3-Month Treasury Bill: Secondary Market Rate",
              yLabel = "Percent" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="DCOILWTICO", Source="FRED", 
              Desc="Crude Oil Prices: West Texas Intermediate (WTI) - Cushing, Oklahoma",
              yLabel = "Dollars per Barrel (Not Adjusted)") )

dfSyms <- rbind(dfSyms, data.frame(Symbol="NEWORDER", Source="FRED", 
              Desc="Manufacturers' New Orders: Nondefense Capital Goods Excluding Aircraft",
              yLabel = "Millions of Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="ALTSALES", Source="FRED", 
              Desc="Light Weight Vehicle Sales: Autos and Light Trucks",
              yLabel = "Millions of Units" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="ICSA", Source="FRED", 
              Desc="Initial Jobless Claims",
              yLabel = "Number" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="^GSPC", Source="yahoo", 
              Desc="S&P 500",
              yLabel = "Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="GDP", Source="FRED", 
              Desc="Gross Domestic Product",
              yLabel = "Billions of Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="GDPC1", Source="FRED", 
              Desc="Real Gross Domestic Product",
              yLabel = "Billions of Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="VIG", Source="yahoo", 
              Desc="Vanguard Dividend Appreciation ETF",
              yLabel = "Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="FEDFUNDS", Source="FRED", 
              Desc="Effective Federal Funds Rate",
              yLabel = "Percent" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="GPDI", Source="FRED", 
              Desc="Gross Private Domestic Investment",
              yLabel = "Billions of Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="MZMV", Source="FRED", 
              Desc=" Velocity of MZM Money Stock",
              yLabel = "Ratio" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="OPHNFB", Source="FRED", 
              Desc="Nonfarm Business Sector: Real Output Per Hour of All Persons",
              yLabel = "Index 2009 = 100" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="IPMAN", Source="FRED", 
              Desc="Industrial Production: Manufacturing (NAICS)",
              yLabel = "Index 2012 = 100" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="^RLG", Source="yahoo", 
              Desc="Russell 1000 Growth ETF",
              yLabel = "Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="IWD", Source="yahoo", 
              Desc="iShares Russell 1000 Value ETF",
              yLabel = "Dollars" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="GS5", Source="FRED", 
              Desc="5-Year Treasury Constant Maturity Rate",
              yLabel = "Percent" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="PSAVERT", Source="FRED", 
              Desc="Personal Saving Rate",
              yLabel = "Percent" ) )

dfSyms <- rbind(dfSyms, data.frame(Symbol="VIXCLS", Source="FRED", 
              Desc="CBOE Volatility Index",
              yLabel = "Index" ) )

```

```{r getsymexample}
# Consumer Price Index for All Urban Consumers: All Items
getSymbols("CPIAUCSL",src="FRED", auto.assign=TRUE)
```

Since it is tedious to do this one at a time, all the symbols were entered into a data frame, loaded, and aggregated together in a single `xts` object. 
This is the complete list of symbol names and sources used in the project.

```{r listsyms, echo=FALSE}
kable(dfSyms)
```

I pull data using quantmod and Quandl. This code pulls in the quantmod data
```{r getsyms, echo=TRUE}
# This pulls in data using quantmod
for (idx in 1:nrow(dfSyms)){
  getSymbols(as.character(dfSyms[idx,"Symbol"]),
             src=as.character(dfSyms[idx,"Source"]), 
             auto.assign=TRUE, 
             from = as.Date("1945-01-01"), to = Sys.Date())
}
```


This adds the Quandl symbols and pulls those in.


```{r getQuandl, echo = TRUE}

# There are a few symbols I wasn't able to include with quantmod so this section uses Quandl to get those data pieces.
# API key 6gea4eLup_eVmnxX9DPA
dfSyms <- rbind(dfSyms, data.frame(Symbol="ISM/MAN_PMI", Source="QUANDL", 
              Desc="Institute of Supply Managment PMI Composite Index",
              yLabel = "Index" ) )

Quandl.api_key('d9LUhcBVPa_8MFdtiFda')
dfPMIComp <- oil_daily <- Quandl("ISM/MAN_PMI", type = "raw", collapse = "daily",  
                    start_date="1910-01-01", end_date=Sys.Date())
strName <- str_replace_all(tail(dfSyms,1)$Symbol, "[^[:alnum:]]", "")
assign(strName,xts(dfPMIComp[,-1], order.by=dfPMIComp[,1]))
colnames(ISMMANPMI) <- c(strName)

strQuandl = "MULTPL/SP500_PE_RATIO_MONTH";
dfSyms <- rbind(dfSyms, data.frame(Symbol=strQuandl, Source="QUANDL", 
              Desc="S&P 500 TTM P/E",
              yLabel = "Index" ) )
dfTemp <- oil_daily <- Quandl(strQuandl, type = "raw", collapse = "daily",  
                                 start_date="1910-01-01", end_date=Sys.Date())

strName <- str_replace_all(tail(dfSyms,1)$Symbol, "[^[:alnum:]]", "")
assign(strName,xts(dfTemp[,-1], order.by=dfTemp[,1]))
colnames(MULTPLSP500PERATIOMONTH) <- c(strName)

strQuandl = "MULTPL/SP500_SALES_QUARTER";
dfSyms <- rbind(dfSyms, data.frame(Symbol=strQuandl, Source="QUANDL", 
              Desc="S&P 500 TTM Sales (Not Inflation Adjusted)",
              yLabel = "Index" ) )
dfTemp <- oil_daily <- Quandl(strQuandl, type = "raw", collapse = "daily",  
                                 start_date="1910-01-01", end_date=Sys.Date())

strName <- str_replace_all(tail(dfSyms,1)$Symbol, "[^[:alnum:]]", "")
assign(strName,xts(dfTemp[,-1], order.by=dfTemp[,1]))
colnames(MULTPLSP500SALESQUARTER) <- c(strName)

strQuandl = "NYXDATA/MARKET_CREDIT";
dfSyms <- rbind(dfSyms, data.frame(Symbol=strQuandl, Source="QUANDL", 
              Desc="NYSE Margin Debt",
              yLabel = "Dollars" ) )
dfTemp <- oil_daily <- Quandl(strQuandl, type = "raw", collapse = "daily",  
                                 start_date="1910-01-01", end_date=Sys.Date())

strName <- str_replace_all(tail(dfSyms,1)$Symbol, "[^[:alnum:]]", "")
assign(strName,xts(dfTemp[,-1], order.by=dfTemp[,1]))
colnames(NYXDATAMARKETCREDIT)[[1]] <- c(strName)


```

This snippet is needed because some of the ticker symbols include are invalid variable names so re-cast those variable names.

```{r symTidy, echo=TRUE}
dfSyms$Symbol <- str_replace_all(dfSyms$Symbol, "[^[:alnum:]]", "")
```

## Feature Extraction

With the raw data downloaded, some of the interesting features can be extracted. The first step is reconcile the time intervals. Some of the data is released monthly and some daily. I chose to interpolate all data to a daily interval. The first section of code adds the daily rows to the dataframe.

```{r aggsyms}
xtsData <- get(dfSyms$Symbol[1])
for(idx in 2:nrow(dfSyms)){
  xtsData <- merge(xtsData,  get(dfSyms$Symbol[idx]) )
  #print(dfSyms$Symbol[idx])
}

dfData <- data.frame(xtsData)

# These two lines evenly space all the data at a daily interval
zooData <- zoo(dfData, as.Date(rownames(dfData)))
dtStartDf <- start(zooData)
dtEndDf <- end(zooData)
zooData <- merge(zooData, zoo(,seq(dtStartDf, dtEndDf, by=1)), all=TRUE)
dfData <- data.frame(zooData)
```

This section here does the interpolation (for continuous) or carries it forward (binary data)

```{r tidydata}
# The recession data is binary and needs to be carried forward
dfData$USREC <- na.locf(dfData$USREC)

dfData$CPIAUCSL <- na.approx(dfData$CPIAUCSL, rule=2)

dfData$UNRATE <- na.approx(dfData$UNRATE, rule=2)
dfData$U6RATE <- na.approx(dfData$U6RATE, rule=2)
dfData$HNONWPDPI <- na.approx(dfData$HNONWPDPI, rule=2)
dfData$GSPC.Open <- na.approx(dfData$GSPC.Open, rule=2)
dfData$GSPC.Volume <- na.approx(dfData$GSPC.Volume, rule=2)
dfData$NOBL.Open <- na.approx(dfData$NOBL.Open, rule=2)
dfData$INDPRO <- na.approx(dfData$INDPRO, rule=2)
dfData$W875RX1A020NBEA <- na.approx(dfData$W875RX1A020NBEA, rule=2)
dfData$RPI <- na.approx(dfData$RPI, rule=2)
dfData$PCOPPUSDM <- na.approx(dfData$PCOPPUSDM, rule=2)
dfData$BUSLOANS <- na.approx(dfData$BUSLOANS, rule =2)
dfData$NCBDBIQ027S <- na.approx(dfData$NCBDBIQ027S, rule = 2 )

dfData$RSALESAGG <- rowMeans(dfData[,c("RRSFS", "RSALES")], na.rm=TRUE)
dfSyms <- rbind(dfSyms, data.frame(Symbol="RSALESAGG", Source="Calc", 
              Desc="Real Retail and Food Services Sales (RRSFS and RSALES)", 
              yLabel ="Millions of Dollars" ) )
dfData$RSALESAGG <- na.approx(dfData$RSALESAGG, rule=2)
dfData$NCBDBIQ027S <- na.approx(dfData$NCBDBIQ027S, rule=2)
dfData$BUSLOANS <- na.approx(dfData$BUSLOANS, rule=2)
dfData$REALLNNSA <- na.approx(dfData$REALLNNSA, rule=2)
dfData$CONSUMERNSA <- na.approx(dfData$CONSUMERNSA, rule = 2)

dfData$TOTLNNSA <- (dfData$BUSLOANS+dfData$REALLNNSA+dfData$CONSUMERNSA)
dfSyms <- rbind(dfSyms, data.frame(Symbol="TOTLNNSA", Source="Calc", 
              Desc="Total Loans Not Seasonally\nAdjusted (BUSLOANS+REALLNSA+CONSUMERNSA)", 
              yLabel ="Billions of U.S. Dollars" ) )

dfData$DGS10 <- na.approx(dfData$DGS10, rule=2)
dfData$DGS1 <- na.approx(dfData$DGS1, rule=2)
dfData$TB3MS <- na.approx(dfData$TB3MS, rule=2)

dfData$DGS10TO1 <- dfData$DGS10-dfData$DGS1
dfSyms <- rbind(dfSyms, data.frame(Symbol="DGS10TO1", Source="Calc", 
              Desc="Yield Curve, 10 and 1 Year Treasury (DGS10-DGS1)", 
              yLabel ="Percent" ) )

dfData$DGS10TOTB3MS <- dfData$DGS10-dfData$TB3MS
dfSyms <- rbind(dfSyms, data.frame(Symbol="DGS10TOTB3MS", Source="Calc", 
              Desc="Yield Curve, 10 and 3 Month Treasury (DGS10-TB3MS)", 
              yLabel ="Percent" ) )

dfData$ALTSALES <- na.approx(dfData$ALTSALES, rule = 2)
dfData$ICSA <- na.approx(dfData$ICSA, rule = 2)
dfData$GDP <- na.approx(dfData$GDP, rule = 2)
dfData$GDPC1 <- na.approx(dfData$GDPC1, rule = 2)
dfData$GPDI <- na.approx(dfData$GPDI, rule = 2)
dfData$MZMV <- na.approx(dfData$MZMV, rule = 2)
dfData$OPHNFB <- na.approx(dfData$OPHNFB, rule = 2)
dfData$DCOILWTICO <- na.approx(dfData$DCOILWTICO, rule = 2)
dfData$ISMMANPMI <- na.approx(dfData$ISMMANPMI, rule=2)
dfData$IPMAN <- na.approx(dfData$IPMAN, rule=2)
dfData$MULTPLSP500PERATIOMONTH <- na.approx(dfData$MULTPLSP500PERATIOMONTH, rule=2)
dfData$MULTPLSP500SALESQUARTER <- na.approx(dfData$MULTPLSP500SALESQUARTER, rule=2)
dfData$FEDFUNDS <- na.approx(dfData$FEDFUNDS, rule=2)
dfData$GS5 <- na.approx(dfData$GS5, rule=2)
dfData$PSAVERT <- na.approx(dfData$PSAVERT, rule=2)
dfData$VIXCLS <- na.approx(dfData$VIXCLS, rule=2)
dfData$NYXDATAMARKETCREDIT <- na.approx(dfData$NYXDATAMARKETCREDIT, rule=2)

```

Year over year and smoothed derivative trends tend to smooth out seasonal variation. It gets used so often that I do this for every series downloaded.

```{r YoYcalcs}

for (strName in names(dfData)){
  strNameYoY <- paste(strName,"_YoY",sep="")
  dfData[strNameYoY] <- CalcYoY(dfData, strName, 365)
  
  strRootSym <- strName
  strSuffix <- ""
  if( grepl("\\.", strName)){
    strRootSym <-  substr(strName, 1, regexpr("\\.", strName)-1)
    strSuffix <- substr(strName, regexpr("\\.", strName)+1, nchar(strName))
    strSuffix <- paste(" (", strSuffix, ")", sep="")
  }
  
  strNewDesc <- dfSyms[grep(paste("^",strRootSym,"$",sep=""), dfSyms$Symbol),]$Desc
  dfSyms <- rbind(dfSyms, data.frame(Symbol=strNameYoY, Source="Calc", 
              Desc=paste(strNewDesc, strSuffix, " Year over Year", sep=""), yLabel="Percent" ))
  
  # Smooth and derivative in one step
  strNewYLabel <- dfSyms[grep(paste("^",strRootSym,"$",sep=""), dfSyms$Symbol),]$yLabel
  strNameSmoothDer <- paste(strName,"_SmoothDer",sep="")
  dfData[strNameSmoothDer] <- sgolayfilt(dfData[,strName], p=3, n=501, m=1, ts=1)
  dfSyms <- rbind(dfSyms, data.frame(Symbol=strNameSmoothDer, Source="Calc", 
              Desc=paste("Derivative of Smoothed ", toTitleCase(strNewDesc), sep=""), 
              yLabel=paste(strNewYLabel, "/period", sep="" ) ) )
  
}
```

These are auxilliary series that use the YoY and smoothed data.

```{r YoYAux}

dfData$GDP_YoYTODGS1 <- dfData$GDP_YoY-dfData$DGS1
dfSyms <- rbind(dfSyms, data.frame(Symbol="GDP_YoYTODGS1", Source="Calc", 
              Desc="Economic Yield Curve, GDP YoY and 1-Year Tr (GDP_YoY-DGS1)", 
              yLabel ="Percent" ) )


dfData$GDP_YoYTOTB3MS <- dfData$GDP_YoY-dfData$TB3MS
dfSyms <- rbind(dfSyms, data.frame(Symbol="GDP_YoYTOTB3MS", Source="Calc", 
              Desc="Economic Yield Curve, GDP YoY and 3-Month Tr (GDP_YoY-TB3MS)", 
              yLabel ="Percent" ) )


dfData$OPHNFB_YoYTODGS1 <- dfData$OPHNFB_YoY-dfData$DGS1
dfSyms <- rbind(dfSyms, data.frame(Symbol="OPHNFB_YoYTODGS1", Source="Calc", 
              Desc="Productivity Yield Curve, Real output YoY and 1-Year Tr (OPHNFB_YoY-DGS1)", 
              yLabel ="Percent" ) )


```

```{r plothelper, echo=FALSE}
# This speeds some of the upcoming operations
dfData$date = as.Date(rownames(dfData))
```

Build the recession and recession initiation dates

```{r recframe}

# The FRED recession data is 1 for the months that
# are in recession and 0 for months not in recession
# so we will use the diff command to save off the 
# indexes where the value changes from 0 to 1 or
# 1 to 0. I found this idea in a stack overflow
# article:
# http://stackoverflow.com/questions/21739012/r-recession-dates-conversion
# I found it was more robust than the nberShade()
# command for the xts time series data.
dtStart <- dfData$date[which(diff(dfData$USREC)==1)+1]
dtEnd   <- dfData$date[which(diff(dfData$USREC)==-1)]
dtInitStart <- as.Date(as.yearmon(dtStart)-12/12)
dtInitEnd <- as.Date(as.yearmon(dtStart)-1/12)

# We need to cast the recession data into
# a dataframe.
dfRecession <- data.frame(initStart = dtInitStart, initEnd = dtInitEnd, start=dtStart, end=dtEnd[-1])
dfRecession <- subset(dfRecession, dtStart >= min(dfData$date))

# Add the recession initiation date as a time series
dfData$RecInit <- rep(0, nrow(dfData))

for( idx in 1:nrow(dfRecession)){
  dfData$RecInit[which(dfData$date>dfRecession$initStart[idx] & dfData$date<dfRecession$initEnd[idx])]=1 
}
dfSyms <- rbind(dfSyms, data.frame(Symbol="RecInit", Source="Calc", 
              Desc="1 for Recession Initiation Period, 0 For All Else", 
              yLabel ="(-)" ) )


dfData$RecInit_Smooth <- sgolayfilt(dfData$RecInit, p=3, n=701, m=0, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="RecInit_Smooth", Source="Calc", 
              Desc="1 for Recession Initiation Period, 0 For All Else (Smoothed)", 
              yLabel ="(-)" ) )


```

Plot the initiation period of each recession

```{r recinit}

datay <- "RecInit"
ylim <- c(-0.1, 1.1)
plotSingleQuick(datay, ylim)

```

Plot the smoothed version of the initiation period of the recession
```{r recinit_smooth}

datay <- "RecInit_Smooth"
ylim <- c(-0.1, 1.1)
plotSingleQuick(datay, ylim)

```


These are auxilliary series that will be used a bit later.

```{r addfeatures}
dfData$UNRATE_Smooth <- sgolayfilt(dfData$UNRATE, p=3, n=21, m=0, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="UNRATE_Smooth", Source="Calc", 
              Desc="Smoothed Civilian Unemployment Rate U-3", 
              yLabel="Percent" ))

dfData$UNRATE_SmoothDer2 <- sgolayfilt(dfData$UNRATE, p=3, n=501, m=2, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="UNRATE_SmoothDer2", Source="Calc", 
              Desc="2nd Derivative of Smoothed U-3", 
              yLabel="Percent/period/period" ))

dfData$U6RATE_Smooth <- sgolayfilt(dfData$U6RATE, p=3, n=21, m=0, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="UNRATE_Smooth", Source="Calc", 
              Desc="Smoothed Total Unemployed U-6", 
              yLabel="Percent" ))

dfData$U6RATE_SmoothDer2 <- sgolayfilt(dfData$U6RATE, p=3, n=501, m=2, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="UNRATE_SmoothDer2", Source="Calc", 
              Desc="2nd Derivative of Smoothed U-6", 
              yLabel="Percent/period/period" ))

dfData$GSPC.OpenLog <- log(dfData$GSPC.Open)
dfSyms <- rbind(dfSyms, data.frame(Symbol="GSPC.OpenLog", Source="Calc", 
              Desc="Log of S&P 500 Open", 
              yLabel="log(Dollar)" ))

dfData$GSPC.VolumeLog <- log(dfData$GSPC.Volume)
dfData$GSPC.VolumeLog[!is.finite(dfData$GSPC.VolumeLog)] <- NA
dfData$GSPC.VolumeLog <- na.approx(dfData$GSPC.VolumeLog, rule=2)
dfSyms <- rbind(dfSyms, data.frame(Symbol="GSPC.VolumeLog", Source="Calc", 
              Desc="Log of S&P 500 Volumne", 
              yLabel="log(shares)" ))

dfData$GSPC.OpenLog_SmoothDer <- sgolayfilt(dfData$GSPC.OpenLog, p=3, n=501, m=1, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="GSPC.OpenLog_SmoothDer", Source="Calc", 
              Desc="Derivative of Smoothed Log Scale S&P 500", 
              yLabel="Dollar/period" ))

dfData$GSPC.OpenLog_SmoothDerDer <- sgolayfilt(dfData$GSPC.OpenLog_SmoothDer, p=3, n=501, m=1, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="GSPC.OpenLog_SmoothDerDer", Source="Calc", 
              Desc="Derivative of Smoothed Log Scale S&P 500", 
              yLabel="Dollar/period/period" ))

dfData$NCBDBIQ027S_Log <- log(dfData$NCBDBIQ027S)
dfSyms <- rbind(dfSyms, data.frame(Symbol="NCBDBIQ027S_Log", Source="Calc", 
              Desc="Log Nonfinancial corporate business;\ndebt securities; liability, Level", 
              yLabel="log(Millions of Dollars)" ))

dfData$NCBDBIQ027S_Log_Der <- sgolayfilt(dfData$NCBDBIQ027S_Log, p=3, n=501, m=1, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="NCBDBIQ027S_Log_Der", Source="Calc", 
              Desc="Derivative of Smoothed Log Nonfinancial corporate business; debt securities; liability, Level", 
              yLabel="log(Milllons of Dollars)/period" ))


dfData$BUSLOANS_Log <- log(dfData$BUSLOANS)
dfSyms <- rbind(dfSyms, data.frame(Symbol="BUSLOANS_Log", Source="Calc", 
              Desc="Log of Consumer and Industrial Loans", 
              yLabel="log(Billlons of U.S. Dollars)" ))

dfData$BUSLOANS_Log_Der <- sgolayfilt(dfData$BUSLOANS_Log, p=3, n=501, m=1, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="BUSLOANS_Log_Der", Source="Calc", 
              Desc="Derivative of Smoothed Log Commercial and Industrial Loans", 
              yLabel="log(Billlons of U.S. Dollars)/period" ))

dfData$GPDI_Log <- log(dfData$GPDI)
dfSyms <- rbind(dfSyms, data.frame(Symbol="GPDI_Log", Source="FRED", 
              Desc="Gross Private Domestic Investment",
              yLabel = "log(Billions of Dollars)" ) )

dfData$GPDI_Log_Der <- sgolayfilt(dfData$GPDI_Log, p=3, n=501, m=1, ts=1)
dfSyms <- rbind(dfSyms, data.frame(Symbol="GPDI_Log_Der", Source="FRED", 
              Desc="Gross Private Domestic Investment",
              yLabel = "log(Billions of Dollars)" ) )


dfData$GDPSP500 <- dfData$GSPC.Close/dfData$GDP
dfData$GDPSP500 <- na.approx(dfData$GDPSP500, rule = 2 )
dfSyms <- rbind(dfSyms, data.frame(Symbol="GDPSP500", Source="Ratio", 
              Desc="S&P 500 (GSPC.Close)/GDP",
              yLabel = "Ratio ($/$)" ) )

dfData$NYXDATAMARKETCREDIT_Log <- log(dfData$NYXDATAMARKETCREDIT)
dfSyms <- rbind(dfSyms, data.frame(Symbol="NYXDATAMARKETCREDIT_Log", Source="QUANDL", 
              Desc="NYSE Margin Debt",
              yLabel = "log(Dollars)" ) )



```

This function calculates the rolling correlation (360-day window), adds it to the dataframe, and plots it out.

```{r rollingcorfun}


calcRollingCorr <- function(dfData, dfSyms, datay1, ylim1, datay2, ylim2, w, dtStart){
  
  # Default for window
  if( missing(w)){
    w <- 360
  }
  
  # Default for the start date for the plot. The rolling correlation is calculated for the entire data range.
  if( missing(dtStart)){
    dtStart <- as.Date("1jan1945","%d%b%Y");
  }
  
  # The calculation
  corrName <- paste(datay1, "_CORR_", datay2, sep="")
  dfData[, corrName] <- RollingCorr(dfData[, datay1], dfData[, datay2], window = w)
  dfData[, corrName][is.na(dfData[, corrName])] <- 1
  
  dfSyms <- rbind(dfSyms, data.frame(Symbol=corrName, Source="calc", 
              Desc=paste("Rolling Correlation ", w, " Day Window"),
              yLabel = "-" ) )
  
  # Return the updated dataframes to the global enviro
  assign('dfData', dfData, envir=.GlobalEnv)
  assign('dfSyms', dfSyms, envir=.GlobalEnv)

  # Plot it out
  p1 <- plotSingleQuickDate(datay1, ylim1, dtStart)
  p2 <- plotSingleQuickDate(datay2, ylim2, dtStart)
  p3 <- plotSingleQuickDate(corrName, c(-1,1), dtStart)

  g1 <- ggplotGrob(p1)
  g2 <- ggplotGrob(p2)
  g3 <- ggplotGrob(p3)
  g <- rbind(g1, g2, g3, size = "first")
  g$widths <- grid::unit.pmax(g1$widths, g2$widths, g3$widths)
  grid.newpage()
  grid.draw(g)
  
  # Return the new name
  return(corrName)

}

```

# Data Exploration

There are some great plotting and visualization tools in the `quantmod` package, but for the prediction work `ggplot` will be used. This section walks through each of the data sets and notes features and relationships that would be helpful in building a trading strategy.

## S&P 500 

One place to begin is with the relationship between stock prices and recessions. The predictor needs to be able to identify the onset of recession before the market declines. In theory, this predictor would give enough warning to cycle out of equity and into something more like cash or bonds. The plot below shows the S&P 500 open values in log-linear format. The market reaches a peak before most recessions, typically 6-9 months before the recession, shown by the blue rectangles in the plot below. A good predictor will correlate with the peak prior to an upcoming recession. 

This section includes the plot code, but since the code is simple it is ommitted for the remainder of the plots. 

```{r SP500plt }
datay <- "GSPC.OpenLog"
ylim <- c(0, 10)
myPlot <- plotSingleQuick(datay, ylim)
myPlot + geom_rect(data=dfRecession,  aes(xmin=initStart, xmax=initEnd, ymin=-Inf, ymax=Inf),
              fill="blue", alpha=0.2, na.rm = TRUE)
```

Take a look at recent activity in the market.

```{r SP500pltNear }
datay <- "GSPC.Open"
ylim <- c(0, 3000)
plotSingleQuickRecentRecent(datay, ylim)

```

The features in the S&P curve that are most interesting are the peaks (where the sell signal should fall) and the troughs (where the buy signal should fall). One way to quantify the peaks is look at the derivative and see where it crosses zero. It looks like most of the positive to negative zero crossing fall in the blue regions where the recession indicator has been defined.

```{r SP500pltder, echo=FALSE}
datay <- "GSPC.OpenLog_SmoothDer"
ylim <- c(-0.005, 0.005)
myPlot <- plotSingleQuick(datay, ylim)
myPlot + geom_rect(data=dfRecession,  aes(xmin=initStart, xmax=initEnd, ymin=-Inf, ymax=Inf),
              fill="blue", alpha=0.2, na.rm = TRUE)
```

The derivative data suggests a way to improve the indicator by adding a single rule: if there is a negative slope zero crossing, move the begin date to thatm point. This approach will allow the algorithm to capture more of the gains of the market. It is also going to be a more challenging fit.

```{r recintmod}

dPeakGSPC <- diff(sign(dfData$GSPC.Open_SmoothDer))
dPeakGSPC <- append(dPeakGSPC, tail(dPeakGSPC,1))
dfData$dPeakGSPC <- dPeakGSPC

# keep only the negative slopes and get rid of any dates
# while the fixed recession indicator is zero
dfData$dPeakGSPC[dfData$dPeakGSPC>0] <- 0
dfData$dPeakGSPC <- (dfData$dPeakGSPC * dfData$RecInit)
dtPeakRec <- dfData$date[which(dfData$dPeakGSPC<0)]

dfRecession$initStartMod <- dfRecession$initStart
for (idxRec in 1: nrow(dfRecession)){
  for( idx in 1:length(dtPeakRec)){
    
    if( dtPeakRec[idx]>dfRecession$initStart[idxRec] & dtPeakRec[idx]<dfRecession$initEnd[idxRec] ){
      
      dfRecession$initStartMod[idxRec] <- dtPeakRec[idx]

    }
    
  }
}

dfRecession$initEndMod <- dfRecession$start + floor((dfRecession$end-dfRecession$start)/2)

# Add the recession initiation date as a time series
dfData$RecInitMod <- rep(0, nrow(dfData))

for( idx in 1:nrow(dfRecession)){
  dfData$RecInitMod[which(dfData$date>dfRecession$initStartMod[idx] & dfData$date<dfRecession$initEndMod[idx])]=1 
}
dfSyms <- rbind(dfSyms, data.frame(Symbol="RecInitMod", Source="Calc", 
              Desc="1 for Recession Initiation Period, 0 For All Else (Modified)", 
              yLabel ="(-)" ) )


datay <- "RecInitMod"
ylim <- c(0.00, 1.00)
myPlot <- plotSingle(dfRecession, dfData, "date", 
                     datay, paste(datay, " | ", dfSyms[grep(paste("^",datay,"$", sep=""), dfSyms$Symbol),]$Desc), 
                     "Date", dfSyms[grep(paste("^",datay,"$", sep=""), dfSyms$Symbol),]$yLabel, 
                     c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, FALSE)
myPlot + geom_rect(data=dfRecession,  aes(xmin=initStartMod, xmax=initEndMod, ymin=-Inf, ymax=Inf),
              fill="green", alpha=0.25, na.rm = TRUE)

datay <- "GSPC.OpenLog_SmoothDer"
ylim <- c(-0.0025, 0.0025)
myPlot <- plotSingle(dfRecession, dfData, "date", 
                     datay, paste(datay, " | ", dfSyms[grep(paste("^",datay,"$", sep=""), dfSyms$Symbol),]$Desc), 
                     "Date", dfSyms[grep(paste("^",datay,"$", sep=""), dfSyms$Symbol),]$yLabel, 
                     c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, FALSE)
myPlot + geom_rect(data=dfRecession,  aes(xmin=initStartMod, xmax=initEndMod, ymin=-Inf, ymax=Inf),
              fill="green", alpha=0.25, na.rm = TRUE)
```


```{r modinit, echo=FALSE}

dfData$G
datay <- "GSPC.OpenLog_SmoothDerDer"
ylim <- c(-0.000025, 0.000025)
myPlot <- plotSingleQuick(datay, ylim)
myPlot + geom_rect(data=dfRecession,  aes(xmin=initStartMod, xmax=initEndMod, ymin=-Inf, ymax=Inf),
              fill="green", alpha=0.25, na.rm = TRUE)

```

The log of the S&P volume has some interesting patterns, but nothing that seems to help with a recession indicator.

```{r SP500Vol, echo=FALSE}

datay <- "GSPC.VolumeLog"
ylim <- c(12, 23)
myPlot <- plotSingleQuick(datay, ylim)
myPlot + geom_rect(data=dfRecession,  aes(xmin=initStart, xmax=initEnd, ymin=-Inf, ymax=Inf),
              fill="blue", alpha=0.2, na.rm = TRUE)

```

## Detrend of Log S&P 500

```{r SP500Vol_demod, echo=FALSE}

model500 <- lm(as.double(dfData$GSPC.VolumeLog) ~ dfData$date)
dfData$GSPC.VolumeLog_demod <- dfData$GSPC.VolumeLog-predict(model500, data.frame(x=dfData$date))
bf <- butter(4, 1/20000, type="high")
dfData$GSPC.VolumeLog_demod <- filtfilt(bf, dfData$GSPC.VolumeLog_demod)
dfData$GSPC.VolumeLog_demod <- filtfilt(bf, dfData$GSPC.VolumeLog)
dfSyms <- rbind(dfSyms, data.frame(Symbol="GSPC.VolumeLog_demod", Source="yahoo", 
              Desc="S&P 500",
              yLabel = "Dollars" ) )


datay <- "GSPC.VolumeLog_demod"
ylim <- c(-2, 2)
plotSingleQuick(datay, ylim)

```

```{r SP500Vol_demod_close, echo=FALSE}

plotSingleQuickRecent(datay, ylim)

```

## S&P 500 TTM P/E

```{r SP500TTMPE }

datay <- "MULTPLSP500PERATIOMONTH"
ylim <- c(0, 100)
plotSingleQuick(datay, ylim)

```

```{r SP500Sales }

datay <- "MULTPLSP500SALESQUARTER"
ylim <- c(0, 1200)
plotSingleQuick(datay, ylim)

```

## NYSE Margin Debt

Taking a look at margin debt. It tends to creep up when there is a frenzy in the stock market.

```{r NYSEMargin }

datay <- "NYXDATAMARKETCREDIT_Log"
ylim <- c(5, 15)
plotSingleQuick(datay, ylim)

```

Sometimes it is more helpful to view year over year growth.

```{r NYSEMargin_Yoy }

datay <- "NYXDATAMARKETCREDIT_YoY"
ylim <- c(-75, 75)
plotSingleQuick(datay, ylim)

```

## CBOE VIX

```{r VIX }

datay <- "VIXCLS"
ylim <- c(0, 45)
plotSingleQuick(datay, ylim)

```

## Dividend Stocks

This is an interesting series, they should perform better through the recessions. Unfortunately they are short lived so there is not much data.

```{r DivStocks }
datay <- "NOBL.Open"
ylim <- c(0, 100)
plotSingleQuick(datay, ylim)

```

## Unemployment rates

Unemployment rates will probably be useful, let's take a look at the U-3. The data is a little noisy so there is also a smoothed version plotted. There seems to be a relationship between the unemployment rate and the recessions, but it could be a lagging indicator.  This will be explored a little bit more later.

```{r unrate, echo=FALSE}
datay <- "UNRATE"
datay_aux <- "UNRATE_Smooth"
ylim <- c(0, 15)
myPlot <- plotSingle(dfRecession, dfData, "date", datay, getPlotTitle(datay), "Date", 
            getPlotYLabel(datay), c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, TRUE)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux, colour=shQuote(datay_aux)), na.rm = TRUE)

```

Looking at the unemployment rate, the eye is drawn to the rise and fall of the data, this suggests that the derivative might be helpful as well. The figure below shows the results, using a Savitzky-Golay FIR filter. It looks like the unemployment rate peaks in the middel of the recession. That peak might be a good buy signal.

```{r UnrateDer, echo=FALSE}

datay = "UNRATE_SmoothDer"
ylim <- c(-0.02, 0.02)
plotSingleQuick(datay, ylim)

```

The second derivative of the unemployment rate does have zero crossings near the middle point of a recession. This would make it a helpful buy signal for the trading strategy.

```{r UnrateDer2, echo=FALSE}

datay = "UNRATE_SmoothDer2"
ylim <- c(-0.0001, 0.0001)
plotSingleQuick(datay, ylim)

```

Let's also take a look at the total unemployed, U-6

```{r unrateU6, echo=FALSE}
datay <- "UNRATE"
datay_aux <- "U6RATE"
ylim <- c(0, 20)
myPlot <- plotSingle(dfRecession, dfData, "date", datay, getPlotTitle(datay), "Date", 
            getPlotYLabel(datay), c(as.Date("1jan1995","%d%b%Y"), Sys.Date()), ylim, TRUE)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux, colour=shQuote(datay_aux)), na.rm = TRUE)

```

## Initial jobless claims

We will also take a look at initial jobless claims, this should start to rise just before the unemployment rate.

```{r initclaims}

datay <- "ICSA"
ylim <- c(100000, 700000)
plotSingleQuick(datay, ylim)

```

It looks like the jobless claim tend to peak more towards the end of the recession. It does not seem to be as strong of a sell indicator as the U-3 rate.

```{r initclaimsDer}
datay <- "ICSA_SmoothDer"
ylim <- c(-1500, 1500)
plotSingleQuick(datay, ylim)

```

Jobless claims have a seasonal component to them. One way to reduce this effect is to calculate year over year growth. That helps some, the peaks seem to be more closely aligned with the middle to end of recessions.

```{r initialclaimsYoy}
datay <- "ICSA_YoY"
ylim <- c(-75, 75)
plotSingleQuick(datay, ylim)

```

## Industrial Production

Industrial production is also known to fall during an economic downturm, let's take a look at some of the data from the FRED on industrual production. It does seem to peak prior to a recession so let's smooth and look at the derivative as it might be a good indicator as well.

```{r indpro, echo=FALSE}
datay <- "INDPRO"
ylim <- c(0, 125)
plotSingleQuick(datay, ylim)

```

The derivative isn't bad, but it sometimes crosses zeros well into a recession. That is less helpful as either a buy or sell indicator. A better measure might year over year (YoY) change.

```{r indproderplot, echo=FALSE}
datay <- "INDPRO_SmoothDer"
ylim <- c(-0.1, 0.1)
plotSingleQuick(datay, ylim)

```

The year over year change has a similar appearance. The low values at the beginning make the year over year values larger than the more recent values. Seems like it will rank low a reliable indicator.

```{r indproYoY, echo=FALSE}
datay <- "INDPRO_YoY"
ylim <- c(-25, 25)
plotSingleQuick(datay, ylim)

```

## Retail Sales

Retail sales also change during recession. As the plot below shows, it seems to follow the trend of industrial production. It might be too strongly correlated to add much to the model. The will be examined in the correlation section.

```{r rsalesagg, echo=FALSE}
datay <- "RSALESAGG"
ylim <- c(0, 200000)
plotSingleQuick(datay, ylim)

```

The derivative of retail sales is a little more erratic than is was the industrial products. Looks like it might be helpful to include in the model as well.

```{r rsalesderplot, echo=FALSE}
datay <- "RSALESAGG_SmoothDer" 
ylim <- c(-100, 100)
plotSingleQuick(datay, ylim)

```

Let's see how that looks on year over year basis
```{r rsalesYoY, echo=FALSE}
datay <- "RSALESAGG_YoY" 
ylim <- c(-15, 15)
plotSingleQuick(datay, ylim)

```

## Industrial production and retail sales correlation

Industrial production and retail sales look very similar so the plot below shows the 360 correlation. The corerlation does tend to fall around a recession, although 2008 was so bad that they both fell together. Not sure if it is that useful.

```{r rollingcortest, fig.width = 10, fig.asp = .62}

datay1 <- "INDPRO"
ylim1 <- c(0, 125)

datay2 <- "RSALESAGG"
ylim2 <- c(0, 200000)

dtStart <- as.Date("1jan1985","%d%b%Y")

w <- 360
corrName <- calcRollingCorr(dfData, dfSyms, datay1, ylim1, datay2, ylim2, w, dtStart)

```

## Real Personal Income

```{r realperinc, echo=FALSE}
datay <- "RPI"
ylim <- c(0, 15000)
plotSingleQuick(datay, ylim)

```

```{r realperincYoY, echo=FALSE}
datay <- "RPI_YoY"
ylim <- c(-10, 10)
plotSingleQuick(datay, ylim)

```

## Consumer price index

What does CPI look like?

```{r CPIPlot}
datay <- "CPIAUCSL"
ylim <- c(0, 300)
plotSingleQuick(datay, ylim)

```

Check out the YoY growth

```{r CPIPlot_YoY}
datay <- "CPIAUCSL_YoY"
ylim <- c(-10, 15)
plotSingleQuick(datay, ylim)

```

## Real Personal Income (Excluding Transfer, Annual)

During a recession real personal income falls. In the plot the peaks can be seen prior to each recession.

```{r realperincex}
datay <- "W875RX1A020NBEA"
ylim <- c(0, 15000)
plotSingleQuick(datay, ylim)

```

The features we are interested in are the peaks and valleys so we'll use the derivative to get to those. Interesting, there is usually a first zero crossing before a recession and a second during or just after the recession.

```{r rperincderplot, echo=FALSE}
datay <- "W875RX1A020NBEA_SmoothDer"
ylim <- c(-2, 2)
plotSingleQuick(datay, ylim)

```

Real personal income might have some seasonal variance, but it seems the year over year change tells the same story.

```{r rperincYoYplotex, echo=FALSE}
datay <- "W875RX1A020NBEA_YoY"
ylim <- c(-10, 10)
plotSingleQuick(datay, ylim)

```

## Copper

Dr. Copper has a reputation as an indicator of economic malaise, but it does not seem to have much of a correlation with the recessions.

```{r cuprice, echo=FALSE}
datay <- "PCOPPUSDM"
ylim <- c(0, 12000)
plotSingleQuick(datay, ylim)

```

I also took a look at the derivative, but it doesn't seem to have much of a correlation either.


```{r coppuderplot, echo=FALSE}
datay <- "PCOPPUSDM_SmoothDer"
ylim <- c(-20, 20)
plotSingleQuick(datay, ylim)

```

## Business Loans 
Business loans should slow before the recession (a contraction in credit as rates rise).

```{r cureal, echo=FALSE}

datay <- "BUSLOANS"
ylim <- c(0, 2500)
plotSingleQuick(datay, ylim)

```

That is just too steep for any kind of machine learnine. This needs to be converted to log scale.

```{r busloanslogplot, echo=FALSE}
datay <- "BUSLOANS_Log"
ylim <- c(0, 10)
plotSingleQuick(datay, ylim)

```

That's a little better, let's see what the smoothed derivative looks like.

```{r busloansderplot, echo=FALSE}
datay <- "BUSLOANS_Log_Der"
ylim <- c(-0.001, 0.001)
plotSingleQuick(datay, ylim)

```

That is odd...looks like this doesn't cross zero unless we are getting close to, or into, a recession. The year over year tells about the same story. 

```{r BusLoansYoYplot, echo=FALSE}

datay <- "BUSLOANS_YoY"
ylim <- c(-30, 30)
plotSingleQuick(datay, ylim)

```

## Nonfinancial Corporate Business Debt

What about Nonfinancial corporate business and debt securities? Hopefully this doesn't follow the business loan trends.

```{r nonfin, echo=FALSE}

datay <- "NCBDBIQ027S"
ylim <- c(-25, 7500000)
plotSingleQuick(datay, ylim)

```

That is crazy steep. Time for a log format, see if that brings out the peaks and troughs. That's a litte better, it looks like there might be a change in slope prior to the recessions.


```{r nonfinlogplot, echo=FALSE}

datay <- "NCBDBIQ027S_Log"
ylim <- c(10, 20)
plotSingleQuick(datay, ylim)

```

The derivative doesn't seem to be much help. There is not much correlation between the zero crossings and the NEBR recessions.

```{r nonfinderplot, echo=FALSE}

datay <- "NCBDBIQ027S_Log_Der"
ylim <- c(-0.0005, 0.0005)
plotSingleQuick(datay, ylim)

```

```{r nonfinderYoYplot, echo=FALSE}

datay <- "NCBDBIQ027S_YoY"
ylim <- c(-30, 30)
plotSingleQuick(datay, ylim)

```

## Total Loans

One business cycle theory describes recessions as a market adjustment to mis-allocated assets, often fueled by an credit expansion. That makes the volume of loans an interesting feature to look at. In the presentation of data it looks like the great recession had the largest impact.

```{r realloans, echo=FALSE}

datay <- "TOTLNNSA"
ylim <- c(-25, 8000)
plotSingleQuick(datay, ylim)

```

Plotting the year over year growth rate helps pull out those small changes in the early years in the data. Peaks can be seen prior to most recessions.

```{r totloans, echo=FALSE}

datay <- "TOTLNNSA_YoY"
ylim <- c(-25, 25)
plotSingleQuick(datay, ylim)

```

```{r totloanssnooth, echo=FALSE}

datay <- "TOTLNNSA_SmoothDer"
ylim <- c(-5, 5)
plotSingleQuick(datay, ylim)

```

As long term interest rates rise, loans should start to tick down. To check this, the total loans and 10 to 1 year spreads are plotted.

```{r totloansYield, echo=FALSE}

datay <- "TOTLNNSA_YoY"
datay_aux<- "DGS10TO1"
ylim <- c(-10, 25)

myPlot <- plotSingle(dfRecession, dfData, "date", datay, getPlotTitle(datay), "Date", 
            getPlotYLabel(datay), c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, TRUE)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux, colour=shQuote(datay_aux)), na.rm = TRUE)


```

There is a good correlation between these two variables. This next section plots that correction explicitly.

## T-Bills and Yield Curve

Speaking of loans, interest rates also play into this. This analysis will focus on treasure bills. The 3-month is plotted below. The yield flattens before a recession as investors go long on bonds and short on equities.

```{r bond3month }

datay <- "TB3MS"
ylim <- c(0, 20)
p1 <- plotSingleQuickModern(datay, ylim)
p1

```

The 1-year is plotted below. The yield flattens before a recession as investors go long on bonds and short on equities.

```{r bond1 }

datay <- "DGS1"
ylim <- c(0, 20)
p1 <- plotSingleQuickModern(datay, ylim)
p1

```

```{r bond10 }

datay <- "DGS10"
ylim <- c(0, 20)
p1 <- plotSingleQuickModern(datay, ylim)
p1

```

Bond yields are a good proxy for interest rates. As rates rise the theory goes that loans should decrease (inverse correlation).

```{r rollingcorLoansInterest, fig.width = 10, fig.asp = .62}

datay1 <- "DGS10"
ylim1 <- c(0, 20)

datay2 <- "TOTLNNSA_YoY"
ylim2 <- c(-25, 25)

dtStart <- as.Date("1jan1985","%d%b%Y")

w <- 360
corrName <- calcRollingCorr(dfData, dfSyms, datay1, ylim1, datay2, ylim2, w, dtStart)

```


The yield curve (10 year bond rate minus the 1 year bond rate) seems to a good indicator of an oncoming recession. It could be a buy indicator by itself.

```{r Yieldcurve, echo=FALSE}
datay <- "DGS10TO1"

ylim <- c(-5, 5)
p1 <- plotSingleQuickModern(datay, ylim)
p1

```

This plot format was suggested by a mises.org article (https://mises.org/wire/yield-curve-accordion-theory), but they only went back to 1988. The date seemed arbitrary so I went back further in time.

```{r YieldAccordian, echo=FALSE}

datay <- "GS5"
datay_aux1<- "DGS1"
datay_aux2<- "DGS10"
datay_aux3 <- "TB3MS"
datay_aux4 <- "FEDFUNDS"
ylim <- c(0, 20)

myPlot <- plotSingle(dfRecession, dfData, "date", datay, getPlotTitle(datay), "Date", 
            getPlotYLabel(datay), c(as.Date("1jan1945","%d%b%Y"), Sys.Date()), ylim, TRUE)
myPlot <- myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux1, colour=shQuote(datay_aux1)), na.rm = TRUE)
myPlot <- myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux2, colour=shQuote(datay_aux2)), na.rm = TRUE)
myPlot <- myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux3, colour=shQuote(datay_aux3)), na.rm = TRUE)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux4, colour=shQuote(datay_aux4)), na.rm = TRUE)

```


Try looking at a 1-year average of the above time series

```{r YieldAccordianYr, echo=FALSE}

GS5Yr <- period.apply(GS5, endpoints(GS5,"years", 1), mean)

```

## Total Loans and Yield Curve Correlation

This relationship was suggest by Charlie and it is an interesting one. As the yield curve flattens (10-year and 1-year rates converge), total loans grow. The generalization is not always accurate, but it does fit.

```{r corrLoansYield, echo=FALSE}

datax = "TOTLNNSA_YoY"
datay = "DGS10TO1"
titlelabel <- paste(datay, " | ", datax)
ylim <- c(-5, 10)
ylabel <- paste(datay, ", ", dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, sep="")
xlim <- c(-10, 20)
xlabel <- paste(datax, ", ", dfSyms[grep(datax, dfSyms$Symbol),]$yLabel, sep="")
bLegend <- FALSE
bFitLinear <- TRUE
dtStart = as.Date('2000-01-01')
plotXvY(dfData, datax, datay, titlelabel, xlabel, ylabel, xlim, ylim, bLegend, bFitLinear, dtStart)

```

## Auto sales

A WSJ article suggested that auto sales might be a good indicator so bring that to the mix. It does have troughs that correlate with recessions

```{r LightAutoSales, echo=FALSE}

datay <- "ALTSALES"
ylim <- c(0, 25)
plotSingleQuick(datay, ylim)

```

There might be some seasonal variance in the auto sales so lets take a look at the year over year. The data is pretty noisy, it probably will not make a very good indicator.

```{r LightAutoSales_YoY, echo=FALSE}

datay <- "ALTSALES_YoY"
ylim <- c(-50, 50)
plotSingleQuick(datay, ylim)

```

## Gross Domestic Product

GDP numbers tend to lag so this series is truly an afterthought. But it does have some correlation with the recessions.

```{r GDP, echo=FALSE}

datay <- "GDP"
ylim <- c(1, 25000)
plotSingleQuick(datay, ylim)


```

Looks like the year over year change on the GDP should correlate well with unemployment.

```{r GDPYoy, echo=FALSE}

datay <- "GDP_YoY"
ylim <- c(-20, 20)
plotSingleQuick(datay, ylim)


```

## Economic Yield Curve

GDP versus the yield on the 1-year. This series was prompted by an article suggesting that the "economic yield curve" should be used to indicate a recession rather than an inverted yield curve.

```{r GDPFundRate, echo=FALSE}

datay <- "GDP_YoY"
datay_aux<- "DGS1"
ylim <- c(-20, 20)

myPlot <- plotSingle(dfRecession, dfData, "date", datay, getPlotTitle(datay), "Date", 
            getPlotYLabel(datay), c(as.Date("1jan1960","%d%b%Y"), Sys.Date()), ylim, TRUE)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux, colour=shQuote(datay_aux)), na.rm = TRUE)

```

```{r GDPFundRateDiff, echo=FALSE}

datay <- "GDP_YoYTODGS1"
ylim <- c(-10.0, 10.0)
plotSingleQuickModern(datay, ylim)


```

```{r GDPFundRate3Month, echo=FALSE}

datay <- "GDP_YoY"
datay_aux<- "TB3MS"
ylim <- c(-20, 20)

myPlot <- plotSingle(dfRecession, dfData, "date", datay, getPlotTitle(datay), "Date", 
            getPlotYLabel(datay), c(as.Date("1jan1960","%d%b%Y"), Sys.Date()), ylim, TRUE)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux, colour=shQuote(datay_aux)), na.rm = TRUE)

```


```{r GDPFundRate3MonthDiff, echo=FALSE}

datay <- "GDP_YoYTOTB3MS"
ylim <- c(-10.0, 10.0)
plotSingleQuickModern(datay, ylim)


```

## Productivity Yield Curve

GDP versus productivity

```{r ProdFundRate, echo=FALSE}

datay <- "OPHNFB_YoY"
datay_aux<- "DGS1"
ylim <- c(-5, 20)

myPlot <- plotSingle(dfRecession, dfData, "date", datay, getPlotTitle(datay), "Date", 
            getPlotYLabel(datay), c(as.Date("1jan1960","%d%b%Y"), Sys.Date()), ylim, TRUE)
myPlot + geom_line(data=dfData, aes_string(x="date", y=datay_aux, colour=shQuote(datay_aux)), na.rm = TRUE)

```
```{r ProdFundRateDiff, echo=FALSE}

datay <- "OPHNFB_YoYTODGS1"
ylim <- c(-20.0, 10.0)
plotSingleQuickModern(datay, ylim)


```

## GDP to S&P 500

GDP to S&P 500

```{r GDPSP500, echo=FALSE}

datay <- "GDPSP500"
ylim <- c(0.00, 0.16)
plotSingleQuick(datay, ylim)


```

## Gross private domestic investment
Spending most certainly tips down prior to a recession. The gross private domestic investment data series, plotted in log format below, show how private investment pulls back prior to recessions.

```{r GPDI_Log, echo=FALSE}

datay <- "GPDI_Log"
ylim <- c(3, 8)
plotSingleQuick(datay, ylim)


```

The change in direction is a little easier to see if the derivative is plotted, first YoY then the smoothed derivative

```{r GPDI_YoY, echo=FALSE}

datay <- "GPDI_YoY"
ylim <- c(-50, 50)
plotSingleQuick(datay, ylim)


```

```{r GPDI_Log_Der, echo=FALSE}

datay <- "GPDI_Log_Der"
ylim <- c(-0.002, 0.002)
plotSingleQuick(datay, ylim)


```

## Velocity

```{r MZMV, echo=FALSE}

datay <- "MZMV"
ylim <- c(1, 4)
plotSingleQuick(datay, ylim)


```

## Personal Savings Rate

Consumers tend to pull down their savings rates as unemployment decreases and market conditions improve.

```{r PSAVERT, echo=FALSE}

datay <- "PSAVERT"
ylim <- c(0, 17.5)
plotSingleQuick(datay, ylim)


```

The relationship between personal savings and unemployment (U-3) can be better visualized with a scatter plot. This was suggested by a WSJ article (https://blogs.wsj.com/dailyshot/2018/02/23/the-daily-shot-reasons-for-declining-u-s-household-savings-rate/).

```{r corrUNPSVT, echo=FALSE}

datax = "UNRATE"
datay = "PSAVERT"
titlelabel <- paste(datay, " | ", datax)
ylim <- c(0, 17.5)
ylabel <- paste(datay, ", ", dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, sep="")
xlim <- c(0, 15)
xlabel <- paste(datax, ", ", dfSyms[grep(datax, dfSyms$Symbol),]$yLabel, sep="")
bLegend <- FALSE
bFitLinear <- TRUE
dtStart = as.Date('2009-01-01')
plotXvY(dfData, datax, datay, titlelabel, xlabel, ylabel, xlim, ylim, bLegend, bFitLinear, dtStart)

```

## Household Net Worth as Percentage of Disposable Personal Income

```{r HNONWPDPI, echo=FALSE}

datay <- "HNONWPDPI"
ylim <- c(450, 700)
plotSingleQuick(datay, ylim)


```


A relationship between personal savings and household networth can be seen in a scatter plot. This was suggested by a WSJ article (https://blogs.wsj.com/dailyshot/2018/02/23/the-daily-shot-reasons-for-declining-u-s-household-savings-rate/).

```{r corrHNONWPDPIPSVT, echo=FALSE}

datax = "HNONWPDPI"
datay = "PSAVERT"
titlelabel <- paste(datay, " | ", datax)
ylim <- c(0, 17.5)
ylabel <- paste(datay, ", ", dfSyms[grep(datay, dfSyms$Symbol),]$yLabel, sep="")
xlim <- c(440, 700)
xlabel <- paste(datax, ", ", dfSyms[grep(datax, dfSyms$Symbol),]$yLabel, sep="")
bLegend <- FALSE
bFitLinear <- TRUE
dtStart = as.Date('1990-01-01')
plotXvY(dfData, datax, datay, titlelabel, xlabel, ylabel, xlim, ylim, bLegend, bFitLinear, dtStart)

```

## Productivity

```{r OPHNFB, echo=FALSE}

datay <- "OPHNFB"
ylim <- c(20, 110)
plotSingleQuick(datay, ylim)


```


```{r OPHNFB_YoY, echo=FALSE}

datay <- "OPHNFB_YoY"
ylim <- c(-10, 10)
plotSingleQuick(datay, ylim)


```

## Crude oil

Look at a trend of West Texas Intermediate (WTI)
```{r Crude, echo=FALSE}

datay <- "DCOILWTICO"
ylim <- c(0, 150)
plotSingleQuick(datay, ylim)


```

```{r Crude_YoY, echo=FALSE}

datay <- "DCOILWTICO_YoY"
ylim <- c(-120, 120)
plotSingleQuick(datay, ylim)


```

## PMI

```{r PMIComp, echo=FALSE}

datay <- "ISMMANPMI"
ylim <- c(30, 90)
plotSingleQuick(datay, ylim)


```

## Industrial Production

This is a look at manufacturing industrial production. The yoY change should be a leading indicator of unemployment.

```{r IPMan, echo=FALSE}

datay <- "IPMAN"
ylim <- c(30, 120)
plotSingleQuick(datay, ylim)


```


```{r IPMan_YoY, echo=FALSE}

datay <- "IPMAN_YoY"
ylim <- c(-25, 25)
plotSingleQuick(datay, ylim)


```

## Correlation Study

Let's see where we are so far. The correlation plot confirms some of the speculation above. The S&P 500 (GSPC.Open) is well correlated with industrial production (INDPRO), business loans (BUSLOANS), total loans (TOTLNNSA) , and nonfinancial corporate business debt (NCBDBIQ027S). 

In this case, I want and indicator that rises prior to a recession. It looks like the unemployment rate (UNRATE), real personal income (W875RX1A020NBEA), and the yield curve (DGS10TO1) are all inversely correlated with the recession initiation indicator.


```{r corplot1, echo=TRUE, fig.width=10,fig.height=11}
# Correlation for the entire data set
  training.cor <- dfData[,c("RecInit","GSPC.Open","GSPC.OpenLog_SmoothDer",
                            "UNRATE","UNRATE_SmoothDer","UNRATE_SmoothDer2", 
                            "INDPRO","INDPRO_SmoothDer","INDPRO_YoY",
                            "RSALESAGG","RSALESAGG_SmoothDer",  
                            "W875RX1A020NBEA","W875RX1A020NBEA_SmoothDer",
                            "W875RX1A020NBEA_YoY",
                            "PCOPPUSDM", "PCOPPUSDM_SmoothDer",
                            "BUSLOANS","BUSLOANS_Log","BUSLOANS_Log_Der","BUSLOANS_YoY",
                            "NCBDBIQ027S", "NCBDBIQ027S_Log","NCBDBIQ027S_Log_Der",
                            "TOTLNNSA","TOTLNNSA_YoY","DGS10TO1", "DGS1", "DGS10",
                            "ALTSALES","ALTSALES_YoY","GDP", "GDPC1", "GDP_YoY",
                            "ICSA","ICSA_YoY", "ICSA_SmoothDer","GPDI","DCOILWTICO", 
                            "GDPSP500","IPMAN")]
  rcorr.data <- rcorr(as.matrix(training.cor), type = "pearson")
  #print(rcorr.data)

  corrplot(cor(training.cor), type="upper", order="original", 
           tl.col="black", tl.srt=45, title ="All data")
```

I thought the modified recession initiation would be a harder match, but there are quite a few correlated variables. Lets take a look at some of those in more detail

# Machine Learning

## Select date range for testing

The data series have different start dates and they are not updated at the same time. This bit of code selects a valid data range to develop the model.
```{r seldata}

mdlFeat <- c("UNRATE", "W875RX1A020NBEA")

dtStart <- index(get(mdlFeat[1])[1]);
for (strSeries in mdlFeat){
  dtThis <- index(get(strSeries)[1])
  if( dtThis > dtStart){
    dtStart <- dtThis
  }
}

dtEnd <- as.Date("2017-01-01")
dfDataModel <- dfData[dfData$date >= dtStart & dfData$date <= dtEnd,]

```

## Partition the data

I break the data into three sets: 50% for training, 25% for testing, and 25% for validation.

```{r partdata, echo=TRUE}
set.seed(123456)
inTrain <- createDataPartition(y=dfDataModel$RecInit, p = 0.50, list=FALSE)
dfTrain <- dfDataModel[inTrain,]
dfData.rest <- dfDataModel[-inTrain,]
inVal <- createDataPartition(y = dfData.rest$RecInit, p = 0.50, list = FALSE)
dfVal <- dfData.rest[inVal,]
dfTest <- dfData.rest[-inVal,]

```


## Create the model

In R, this is the easiest part. I favor simple models because they tend to be more robust. From the data exploration I know that the yield curve is variable with respect to recessions and my recession initiation indicator. I am going to start with the unemployment rate and personal income and see how the model performs.

```{r treepart, echo=TRUE}
strEqnPart <- paste("RecInit ~ ", paste(mdlFeat, collapse = "+", sep=""))
modFit <- rpart( strEqnPart, data=dfTrain)
```

The plot below summarizes the correlations in a graphical format. The tree itself is a little more complicated than I like.

```{r treeplot, echo=TRUE, fig.width=10,fig.height=11}
rpart.plot(modFit, main="Recession Indicator", type = 2, extra = 1)
```


## Validation

Rather than step through a quantitative validation I am going going to plot the data against the NEBR recessions and the recession initiation indicator. There is good agreement between the model and indicator. We can create a buy signal anytime that the indicator is above 0.5

```{r predplot}

dfData$RecInitPred <- predict(modFit, newdata=dfData)
dfSyms <- rbind(dfSyms, data.frame(Symbol="RecInitPred", Source="Calc", 
              Desc="Prediction 1 for Recession Initiation Period, 0 For All Else", 
              yLabel ="(-)" ) )


datay <- "RecInitPred"
datay_aux<- "RecInit"
ylim <- c(0, 1)
myPlot <- plotSingleQuickDate(datay, ylim, dtStart)
myPlot + geom_rect(data=dfRecession,  aes(xmin=initStart, xmax=initEnd, ymin=-Inf, ymax=Inf),
              fill="blue", alpha=0.2, na.rm = TRUE)


```

# Creating the Trading Rules 

The strategy here will be to go long when the recesion initiation signal is below 0.5. When it crosses 0.5 I will exit the market. I need a signal that will tell me when to get back in. From the data exploration section I noted that the second derivative of the unemployment rate crosses zero right in the middle of most recessions That zero crossing will serve as the buy signal to get back in the market.

## Create the trade rule for recession initiation

```{r buysig, echo=TRUE}

# Round off the prediction
dfData$RecInitPredRd <- round(dfData$RecInitPred, digits = 0)
dfSyms <- rbind(dfSyms, data.frame(Symbol="RecInitPredRd", Source="Calc", 
              Desc="Rounded Prediction 1 for Recession Initiation\n Period, 0 For All Else", 
              yLabel ="(-)" ) )

# Extract the dates from the prediction
dtStartPred <- dfData$date[which(diff(dfData$RecInitPredRd)==1)+1]
dtStartPred <- dtStartPred+30
dtEndPred <- dfData$date[which(diff(dfData$RecInitPredRd)==-1)]

# Zero crossings of the 2nd derivative of the unemployment rate
dZeroUNRATE <- diff(sign(dfData$UNRATE_SmoothDer2))
dZeroUNRATE <- append(dZeroUNRATE, tail(dZeroUNRATE,1))
dfData$dZeroUNRATE <- dZeroUNRATE

# keep only the negative slopes and get rid of any dates
# while the indicator is high
dfData$dZeroUNRATE[dfData$dZeroUNRATE>0] <- 0
dfData$dZeroUNRATE <- -1 * (dfData$dZeroUNRATE * (dfData$RecInitPredRd - 1))


# Find the first negative slope after the recession indicator triggers
dtEndCand <- dfData$date[which(dfData$dZeroUNRATE<0)]
dtBuyPred <- dtStartPred
for (idx in 1:length(dtStartPred)){
  dtBuyPred[idx] <- dtEndCand[min(which(dtEndCand > dtEndPred[idx]))]
}
dfPred <- data.frame(predStart = dtStartPred, predEnd = dtBuyPred)

# Create the trade rule, 1 is long, 0 is not invested
dfData$RecInitTrade <- rep(1, nrow(dfData))

for( idx in 1:nrow(dfPred)){
  dfData$RecInitTrade[which(dfData$date>dfPred$predStart[idx] & dfData$date<dfPred$predEnd[idx])]=0 
}
dfSyms <- rbind(dfSyms, data.frame(Symbol="RecInitTrade", Source="Calc", 
              Desc="Recession Initiation Trade Rule", 
              yLabel ="(-)" ) )

```


```{r backval, include=FALSE}
# Create a data series and validate that our back testing function works correctly.
# Create a 12-month series with 10% annual interest compounded annually
dTest <- UNRATE['1999/2001']
dRate <- 0.1
dTest[,'UNRATE'] = ((1+dRate/12.0) ^ (12.0*seq(1:nrow(dTest))/12))
dMonths <- 24

# Test case 1, 100% in for the whole time
dEqAn <- (1+dRate/12.0)^(12.0*(dMonths/12.0))
ret <- ROC(dTest)
ret <- ret['2000/2001']
dEq1Nm <- exp(cumsum(ret))
dEq1Nm <- as.numeric(tail(dEq1Nm,1)$UNRATE)

dfTestResults <- data.frame(c("100%"), c(dEqAn), c(dEq1Nm))
colnames(dfTestResults) <- c("Test","Analytical", "Numerical")
dfTestResults

```

## Plot the Backtesting Results

### Create the Baseline Data Series

The trading strategy will be compared to the S&P 500, shown below. We use this to create an S&P 500 rate of change series. The trading rule will move in and out of this series.

```{r plotbackbase }

dfData$retBase <- ROC(dfData$GSPC.Close)
dfData$retBase[is.na(dfData$retBase)] <- 0
dfSyms <- rbind(dfSyms, data.frame(Symbol="retBase", Source="Calc", 
              Desc="S&P 500 Rate of Change", 
              yLabel ="Percent" ) )

dfData$eqBase <- exp(cumsum(dfData$retBase))
dfSyms <- rbind(dfSyms, data.frame(Symbol="eqBase", Source="Calc", 
              Desc="Equity Return, 100% long", 
              yLabel ="Percent" ) )

datay <- "eqBase"
ylim <- c(0, 1000)
p1 <- plotSingleQuick(datay, ylim)

datay <- "retBase"
ylim <- c(-0.1, 0.1)
p2 <- plotSingleQuick(datay, ylim)

grid.arrange(p1, p2, ncol = 1, top = paste("Base Case (100% long) | Growth = ", sprintf('%0.2f', tail(dfData$eqBase,1)), sep=""))

```

### Perform the Backtesting and Plot the Results

In this final analysis step the trading rule is plotted along with the indicator in the top pane. The middle pane shows how the trading rule modified the rate of change series. The bottom plots shows how the investment performed, compared to the S&P 500. 

```{r plotbackrecInit,fig.width=7,fig.height=10}

dfData$retRec <- dfData$retBase * dfData$RecInitTrade
dfSyms <- rbind(dfSyms, data.frame(Symbol="retRec", Source="Calc", 
              Desc="Rate of Change, Recession Initiation Rule", 
              yLabel ="Percent" ) )

dfData$eqRec <- exp(cumsum(dfData$retRec))
dfSyms <- rbind(dfSyms, data.frame(Symbol="eqRec", Source="Calc", 
              Desc="Equity Return, Recession Initiation Rule", 
              yLabel ="Percent" ) )

dataTrade <- "RecInitTrade"
dataRet <- "retRec"
dataEq <- "eqRec"
p1 <- plotBack(dataTrade, dataRet, dataEq, dfPred)

```

The trading strategy resulted in an improved return, although most of this comes after the 2007-2008 recession. In that recession the trading rules return to the long position at the exact market bottom. This is in contrast to the 1999-2000 recession where the trading rule returns to long before the market bottoms.

# Conclusion

In this worksheet a model predicting the onset of recession was built. From the model a trading rule was derived to allow backtesting. The model performed well and the trading rule backtesting showed that applying this in the post-WWII period would have resulted in a 14.5% increase in returns. That is not too bad, but there are a few changes that would likely improve the model:

- Go long on short term bonds, rather than just roll out of the market. That way at least some returns would be generated during recessions.
- Refine the recession indicator. The square shape of the indicator makes it difficult to model. Something more like a windowed sync function might give better results.

As it is the model does provide value. It will be interesting to see how it hold up over the next few years.


